### 2015-08-25 article/ms.tex
Article init

\documentclass[useAMS,usenatbib]{mn2e}

\newcommand\tc{\textit{The Cannon}}
\title[Cannon Chemistry]{Extending \tc{} to Individual Chemical Abundances} 
% Some arrangement of the following authors:
\author[Casey et al.]{A.~R.~Casey$^1$, M.~Ness$^2$, G.~Gilmore$^1$, D.~W.~Hogg$^{2,3,4}$, H.~W.~Rix$^2$ \\ 
$^1$Institute of Astronomy, University of Cambridge, Madingley Road, Cambridge
    CB3 0HA, UK\\
$^2$Max-Planck-Institut f\"ur Astronomie, K\"onigstuhl 17, D-69117 Heidelberg,
    Germany\\
$^3$Center for Cosmology and Particle Physics, Department of Physics, New York
    University, 4 Washington Pl., room 424, New York, \\
    NY, 10003, USA\\
$^4$Center for Data Science, New York University, 726 Broadway, 7th Floor,
    New York, NY 10003, USA}
\begin{document}

\date{Accepted 2015 XX XX. Received 2015 October XX; in original form 2015 XX xx}

\pagerange{\pageref{firstpage}--\pageref{lastpage}} \pubyear{2015}

\maketitle

\label{firstpage}

\begin{abstract}
\end{abstract}

\begin{keywords}
\end{keywords}

\section{Introduction}

\section{Conclusions}

\section*{Acknowledgments}
This research was fostered by discussions at the \textit{New Milky Way} conference hosted by the Munich Institute for Astro- and Particle Physics, and the \textit{Stellar Streams in the Local Universe} conference hosted at Ringberg Castle by the Max-Planck-Institut F\"ur Astronomie. This research made use of Astropy, a community-developed core Python package for Astronomy \citep{astropy}.

\begin{thebibliography}{99}
%\bibitem[\protect\citeauthoryear{Baird}{1981}]{b1} Baird S.R., 1981,
%ApJ, 245, 208
\bibitem[Astropy Collaboration et al.(2013)]{astropy} Astropy Collaboration, Robitaille, T.~P., Tollerud, E.~J., et al.\ 2013, \aap, 558, AA33
\end{thebibliography}

\label{lastpage}

\end{document}

### 2015-08-25 explosives/__init__.py
Update logging and directory structure

@@ -10,8 +10,11 @@ import logging
 from numpy import RankWarning
 from warnings import simplefilter
 
-logging.basicConfig(level=logging.INFO, 
-    format="%(asctime)s [%(levelname)s] %(message)s")
+logger = logging.getLogger("explosives")
+logger.setLevel(logging.INFO)
+
+#logging.basicConfig(level=logging.INFO, 
+#    format="%(asctime)s [%(levelname)s] %(message)s")
 
 simplefilter("ignore", RankWarning)
 simplefilter("ignore", RuntimeWarning)

### 2015-08-25 explosives/cannon.py

@@ -12,7 +12,7 @@ import numpy as np
 import scipy.optimize as op
 from . import (model, plot, utils)
 
-logger = logging.getLogger("cannon")
+logger = logging.getLogger("explosives")
 
 
 class CannonModel(model.BaseModel):

### 2015-08-25 explosives/cannon.py
Optionally save/load models with the data that was used to train it

@@ -518,7 +518,7 @@ class CannonModel(model.BaseModel):
 
 
     @model.requires_training_wheels
-    def save(self, filename, overwrite=False, verify=True):
+    def save(self, filename, with_data=False, overwrite=False, verify=True):
         """
         Save the (trained) model to disk. This will save the label vector
         description, the optimised coefficients and scatter, and pivot offsets.
@@ -528,6 +528,13 @@ class CannonModel(model.BaseModel):
 
         :type filename:
             str
+
+        :param with_data: [optional]
+            Save the wavelengths, fluxes and flux uncertainties used to train
+            the model.
+
+        :type with_data:
+            bool
 
         :param overwrite: [optional]
             Overwrite the existing file path, if it already exists.
@@ -550,8 +557,12 @@ class CannonModel(model.BaseModel):
             hashes = None
 
         contents = \
-            (self._label_vector_description, self._coefficients, self._scatter,
-                self._offsets, hashes)
+            [self._label_vector_description, self._coefficients, self._scatter,
+                self._offsets, hashes]
+        if with_data:
+            contents.extend(
+                [self._wavelengths, self._fluxes, self._flux_uncertainties])
+
         with open(filename, "w") as fp:
             pickle.dump(contents, fp, -1)
 
@@ -601,8 +612,13 @@ class CannonModel(model.BaseModel):
                         "to that stored in {2} ({3})".format(descr, e_hash,
                             filename, r_hash)) 
 
-        self._label_vector_description, self._coefficients, self._scatter, \
-            self._offsets, hashes = contents
+        if len(contents) > 5:
+            self._label_vector_description, self._coefficients, self._scatter, \
+                self._offsets, hashes, self._wavelengths, self._fluxes, \
+                self._flux_uncertainties = contents
+        else:
+            self._label_vector_description, self._coefficients, self._scatter, \
+                self._offsets, hashes = contents
         self._trained = True
 
         return True

### 2015-08-26 explosives/atomic.py
Basic EW->log(X) model

@@ -1,0 +1,312 @@
+#!/usr/bin/env python
+# -*- coding: utf-8 -*-
+
+""" Approximately model an atomic absorption line in a stellar spectrum. """
+
+__author__ = "Andy Casey <arc@ast.cam.ac.uk>"
+
+import logging
+import numpy as np
+import scipy.optimize as op
+
+import oracle.synthesis
+
+logger = logging.getLogger("explosives")
+
+
+def approximate_atomic_transitions(stellar_parameters, transitions,
+    ew_points=None, photosphere_kwargs=None, synthesis_kwargs=None, **kwargs):
+    """
+    Return functions to approximate the equivalent width of atomic transitions
+    given the stellar parameters and logarithmic abundance.
+
+    :param stellar_parameters:
+        A multi-row table containing the temperature, surface gravity,
+        metallicity(, and microturbulence) for stars in the training set. If the
+        microturbulence is not provided, it will be approximated with the Kirby
+        et al. (2008) relation for giants (logg < 3.5) and from the Reddy et al. 
+        (2003) relationship for dwarfs (logg >= 3.5).
+
+    :type stellar_parameters:
+        :class:`np.ndarray`
+
+    :param transitions:
+        A table containing the atomic data for each transition to build a model
+        for. This table should include columns named: `wavelength`, `species`,
+        `excitation_potential`, `log_gf`, and optionally: `VDW_DAMP`.
+
+    :type transitions:
+        :class:`~astropy.table.Table`
+
+    :param ew_points: [optional]
+        The equivalent widths (EWs) to sample at each set of stellar parameters
+        for each atomic line. If `None` is provided, then 30 EW values between
+        (1, 300) milliAngstroms will be sampled for each star.
+
+    :type ew_points:
+        list
+
+    :param photosphere_kwargs: [optional]
+        Keywords that will be used when generating photospheres.
+
+    :type photosphere_kwargs:
+        dict
+
+    :param synthesis_kwargs: [optional]
+        Keywords that will be used when running MOOG to determine abundances.
+
+    :type synthesis_kwargs:
+        dict
+    """
+
+    import cPickle as pickle
+
+    if False:
+        # Calculate abundances for all lines in all stars.
[…]

### 2015-08-26 explosives/cannon.py

@@ -10,7 +10,7 @@ import logging
 import numpy as np
 
 import scipy.optimize as op
-from . import (model, plot, utils)
+from . import (atomic, model, plot, utils)
 
 logger = logging.getLogger("explosives")
 
@@ -63,7 +63,7 @@ class CannonModel(model.BaseModel):
             The human-readable form of the label vector description.
 
         :type label_vector_description:
-            str
+            str or list of str
 
         :param N: [optional]
             Limit the number of stars used in the training set. If left to None,

### 2015-08-26 explosives/cannon.py
Building frameworks [ci skip]

@@ -10,7 +10,7 @@ import logging
 import numpy as np
 
 import scipy.optimize as op
-from . import (atomic, model, plot, utils)
+from . import (model, plot, utils)
 
 logger = logging.getLogger("explosives")
 
@@ -50,8 +50,8 @@ class CannonModel(model.BaseModel):
             :class:`np.ndarray`
         """
 
-        super(self.__class__, self).__init__(labels, fluxes, flux_uncertainties,
-            verify=verify, wavelengths=wavelengths)
+        super(CannonModel, self).__init__(labels, fluxes, flux_uncertainties,
+            wavelengths, verify)
 
 
     def train(self, label_vector_description, N=None, limits=None, pivot=False,
@@ -103,8 +103,8 @@ class CannonModel(model.BaseModel):
         coefficients = np.nan * np.ones((N_pixels, lva.shape[0]))
         
         pb_size = 100 if __kwargs.pop("__progressbar", True) else 0
-        pb_message = "Training Cannon model from {0} stars with {1} pixels:\n"\
-            .format(N_stars, N_pixels)
+        pb_message = "Training {0} model from {1} stars with {2} pixels:\n"\
+            .format(self.__class__.__name__[:-5], N_stars, N_pixels)
         for i in utils.progressbar(range(N_pixels), pb_message, pb_size):
 
             coefficients[i, :], scatter[i] = _fit_pixel(
@@ -476,7 +476,7 @@ class CannonModel(model.BaseModel):
         return theta
 
 
-    def _repr_label_vector_description(self, label_vector_indices):
+    def _repr_label_vector_description(self, label_vector_indices, **kwargs):
         """
         Represent label vector indices as a readable label vector description.
 
@@ -488,7 +488,7 @@ class CannonModel(model.BaseModel):
             A human-readable string of the label vector description.
         """
 
-        string = ["1"]
+        string = [kwargs.pop("__first_vector_desc", "1")]
         for cross_terms in label_vector_indices:
             sub_string = []
             for descr, order in cross_terms:

### 2015-08-26 explosives/explosives.py

@@ -1,0 +1,508 @@
+#!/usr/bin/env python
+# -*- coding: utf-8 -*-
+
+""" Cannon with Chemistry """
+
+__author__ = "Andy Casey <arc@ast.cam.ac.uk>"
+
+import cPickle as pickle
+import logging
+import numpy as np
+
+import scipy.optimize as op
+from . import (model, cannon, plot, utils)
+
+logger = logging.getLogger("explosives")
+
+
+class ExplosivesModel(cannon.CannonModel):
+
+    def __init__(self, labels, wavelengths, fluxes, flux_uncertainties,
+        verify=True):
+        """
+        Initialise a Cannon-with-Chemistry model.
+
+        :param labels:
+            A table with columns as labels, and stars as rows.
+
+        :type labels:
+            :class:`~astropy.table.Table`
+
+        :param wavelengths:
+            The wavelengths of the given pixels.
+
+        :type wavelengths:
+            :class:`np.array`
+
+        :param fluxes:
+            An array of fluxes for each star as shape (num_stars, num_pixels).
+            The num_stars should match the rows in `labels`.
+
+        :type fluxes:
+            :class:`np.ndarray`
+
+        :param flux_uncertainties:
+            An array of 1-sigma flux uncertainties for each star as shape
+            (num_stars, num_pixels). The shape of the `flux_uncertainties` array
+            should match the `fluxes` array. 
+
+        :type flux_uncertainties:
+            :class:`np.ndarray`
+        """
+
+        super(self.__class__, self).__init__(labels, wavelengths, fluxes,
+            flux_uncertainties, verify)
+
+
+    def train(self, label_vector_description, atomic_lines=None, N=None,
+        limits=None, pivot=False, **kwargs):
+        """
+        Train a Cannon model based on the label vector description provided.
+
+        :params label_vector_description:
+            The human-readable form of the label vector description.
+
+        :type label_vector_description:
+            str or list of str
+
+        :params atomic_lines: [optional]
+            Atomic absorption lines that should be modelled as part of the first
+            entry in the label vector. If given, this should be a dictionary
+            containing the label names that correspond to $\log_\epsilon(X)$
+            abundances in `CannonModel._labels` as keys, and the values should
+            be an `~astropy.table.Table` containing the atomic transitions for
+            that element.
+
+        :type atomic_lines:
+            dict
+
+        :param N: [optional]
+            Limit the number of stars used in the training set. If left to None,
+            all stars will be used.
+
+        :type N:
+            None or int
+
+        :param limits: [optional]
[…]

### 2015-08-27 explosives/atomic.py
Refactored a little, improved code

@@ -9,12 +9,12 @@ import logging
 import numpy as np
 import scipy.optimize as op
 
-import oracle.synthesis
+from oracle import photospheres, synthesis
 
 logger = logging.getLogger("explosives")
 
 
-def approximate_atomic_transitions(stellar_parameters, transitions,
+def approximate_atomic_transitions(stellar_parameters, transitions, X_H=False,
     ew_points=None, photosphere_kwargs=None, synthesis_kwargs=None, **kwargs):
     """
     Return functions to approximate the equivalent width of atomic transitions
@@ -38,10 +38,17 @@ def approximate_atomic_transitions(stellar_parameters, transitions,
     :type transitions:
         :class:`~astropy.table.Table`
 
+    :param X_H: [optional]
+        Use abundances in X_H format. If set to `False`, then log(X) abundance
+        formats are assumed.
+
+    :type X_H:
+        bool
+
     :param ew_points: [optional]
         The equivalent widths (EWs) to sample at each set of stellar parameters
         for each atomic line. If `None` is provided, then 30 EW values between
-        (1, 300) milliAngstroms will be sampled for each star.
+        (10, 300) milliAngstroms will be sampled for each star.
 
     :type ew_points:
         list
@@ -64,7 +71,7 @@ def approximate_atomic_transitions(stellar_parameters, transitions,
     if False:
         # Calculate abundances for all lines in all stars.
         ews, abundances = _atomic_line_abundances(stellar_parameters, transitions,
-            ew_points=ew_points, photosphere_kwargs=photosphere_kwargs,
+            X_H=X_H, ew_points=ew_points, photosphere_kwargs=photosphere_kwargs,
             synthesis_kwargs=synthesis_kwargs, **kwargs)
 
         with open("tmp.pkl", "wb") as fp:
@@ -72,6 +79,10 @@ def approximate_atomic_transitions(stellar_parameters, transitions,
     else:
         with open("tmp.pkl", "rb") as fp:
             ews, abundances = pickle.load(fp)
+
+        if X_H:
+            logger.warn("ASSUMING ALL FE")
+            abundances -= 7.5
 
     # f(teff, logg, EW) --> logX
[…]

### 2015-08-27 explosives/cannon.py
Refactored save/load to basemodel

@@ -17,6 +17,9 @@ logger = logging.getLogger("explosives")
 
 class CannonModel(model.BaseModel):
 
+    _trained_attributes = ("_coefficients", "_scatter", "_offsets")
+    _data_attributes = ("_wavelengths", "_fluxes", "_flux_uncertainties")
+
     def __init__(self, labels, wavelengths, fluxes, flux_uncertainties,
         verify=True):
         """
@@ -503,125 +506,6 @@ class CannonModel(model.BaseModel):
                 string.append(sub_string[0])
 
         return " + ".join(string)
-
-
-    @property
-    def _trained_hash(self):
-        """
-        Return a hash of the trained state.
-        """
-
-        if not self._trained: return None
-        args = (self._coefficients, self._scatter, self._offsets,
-            self._label_vector_description)
-        return "".join([str(hash(str(each)))[:10] for each in args])
-
-
-    @model.requires_training_wheels
-    def save(self, filename, with_data=False, overwrite=False, verify=True):
-        """
-        Save the (trained) model to disk. This will save the label vector
-        description, the optimised coefficients and scatter, and pivot offsets.
-
-        :param filename:
-            The file path where to save the model to.
-
-        :type filename:
-            str
-
-        :param with_data: [optional]
-            Save the wavelengths, fluxes and flux uncertainties used to train
-            the model.
-
-        :type with_data:
-            bool
-
-        :param overwrite: [optional]
-            Overwrite the existing file path, if it already exists.
-
-        :type overwrite:
-            bool
-
-        :returns:
-            True
-
-        :raise TypeError:
-            If the model has not been trained, since there is nothing to save.
-        """
-
-        # Create a hash of the labels, fluxes and flux uncertainties.
-        if verify:
-            hashes = [hash(str(_)) for _ in \
-                (self._labels, self._fluxes, self._flux_uncertainties)]
-        else:
-            hashes = None
-
-        contents = \
-            [self._label_vector_description, self._coefficients, self._scatter,
-                self._offsets, hashes]
-        if with_data:
-            contents.extend(
-                [self._wavelengths, self._fluxes, self._flux_uncertainties])
-
-        with open(filename, "w") as fp:
-            pickle.dump(contents, fp, -1)
-
-        return True
-
-
-    def load(self, filename, verify=True):
-        """
-        Load a trained model from disk.
-
-        :param filename:
-            The file path where to load the model from.
-
-        :type filename:
-            str
-
-        :param verify: [optional]
[…]

### 2015-08-27 explosives/model.py

@@ -18,8 +18,16 @@ def requires_training_wheels(f):
     return wrapper
 
 
+def _short_hash(self, contents):
+    """ Return a short hash of some iterable contents. """
+    return "".join([str(hash(str(item)))[:10] for item in contents])
+
+
 class BaseModel(object):
 
+    _trained_attributes = None
+    _data_attributes = None
+    
     def __init__(self, labels, fluxes, flux_uncertainties, wavelengths=None,
         verify=True):
         """
@@ -145,4 +153,123 @@ class BaseModel(object):
                     raise ValueError("forbidden character '{0}' is in potential"
                         " label '{1}' - to ignore this use verify=False".format(
                             character, column))
+        return True
+
+
+    @property
+    def _trained_hash(self):
+        """ Return a hash of the trained state. """
+
+        if not self._trained: return None
+        return _short_hash([getattr(self, _) for _ in self._trained_attributes])
+
+
+    @property
+    def _data_hash(self):
+        """ Return a hash of the data. """
+        return _short_hash([getattr(self, _) for _ in self._data_attributes])
+
+
+    @model.requires_training_wheels
+    def save(self, filename, with_data=False, overwrite=False):
+        """
+        Save the (trained) model to disk. This will save the label vector
+        description, the optimised coefficients and scatter, and pivot offsets.
+
+        :param filename:
+            The file path where to save the model to.
+
+        :type filename:
+            str
+
+        :param with_data: [optional]
+            Save the wavelengths, fluxes and flux uncertainties used to train
+            the model.
+
+        :type with_data:
+            bool
+
+        :param overwrite: [optional]
+            Overwrite the existing file path, if it already exists.
+
+        :type overwrite:
+            bool
+
+        :returns:
+            True
+
+        :raise TypeError:
+            If the model has not been trained, since there is nothing to save.
+        """
+
+        assert None not in (self._data_attributes, self._trained_attributes)
+
+        contents = [getattr(self, _) for _ in self._trained_attributes]
+        contents += [self._data_hash]
+        if with_data:
+            contents.extend([getattr(self, _) for _ in self._data_attributes])
+
+        with open(filename, "w") as fp:
+            pickle.dump(contents, fp, -1)
+
+        return True
+
+
+    def load(self, filename, verify=True):
+        """
+        Load a trained model from disk.
+
+        :param filename:
+            The file path where to load the model from.
+
+        :type filename:
+            str
+
+        :param verify: [optional]
+            Verify whether the hashes in the stored filename match what is
[…]

### 2015-08-27 explosives/explosives.py
Progress made in weak line integration

@@ -10,12 +10,15 @@ import logging
 import numpy as np
 
 import scipy.optimize as op
-from . import (model, cannon, plot, utils)
+from . import (atomic, cannon, model, plot, utils)
 
 logger = logging.getLogger("explosives")
 
 
 class ExplosivesModel(cannon.CannonModel):
+
+    _save_attributes = ("_coefficients", "_scatter", "_offsets", "_ew_model")
+    _data_attributes = ("_wavelengths", "_fluxes", "_flux_uncertainties")
 
     def __init__(self, labels, wavelengths, fluxes, flux_uncertainties,
         verify=True):
@@ -54,8 +57,8 @@ class ExplosivesModel(cannon.CannonModel):
             flux_uncertainties, verify)
 
 
-    def train(self, label_vector_description, atomic_lines=None, N=None,
-        limits=None, pivot=False, **kwargs):
+    def train(self, label_vector_description, atomic_lines=None, X_H=False,
+        N=None, limits=None, pivot=False, **kwargs):
         """
         Train a Cannon model based on the label vector description provided.
 
@@ -75,6 +78,13 @@ class ExplosivesModel(cannon.CannonModel):
 
         :type atomic_lines:
             dict
+
+        :param X_H: [optional]
+            Use abundances in X_H format. If set to `False`, then log(X)
+            abundance formats are assumed.
+
+        :type X_H:
+            bool
 
         :param N: [optional]
             Limit the number of stars used in the training set. If left to None,
@@ -116,23 +126,83 @@ class ExplosivesModel(cannon.CannonModel):
         lva, use, offsets = cannon._build_label_vector_array(self._labels, lv,
             N, limits, pivot)
 
+        # Initialise the requisite arrays.
+        N_stars, N_pixels = self._fluxes.shape[:2]
+        scatter = np.nan * np.ones(N_pixels)
+        coefficients = np.nan * np.ones((N_pixels, lva.shape[0]))
+        weak_line_fluxes = np.ones((N_stars, N_pixels))
+        
         # Any atomic lines to model?
-        p_sigma = -1
-        valid_atomic_lines = _check_atomic_lines(self._labels, atomic_lines)
-        if valid_atomic_lines:
+        if _check_atomic_lines(self._labels, atomic_lines):
+
+            N_species = len(atomic_lines)
+            N_transitions = sum(map(len, atomic_lines.values()))
+            logger.info("Including {0} weak lines from {1} elements: {2}".format(
[…]

### 2015-08-27 explosives/cannon.py
Some fixes, and models can be loaded by initialisation or from existing model class

@@ -5,7 +5,6 @@
 
 __author__ = "Andy Casey <arc@ast.cam.ac.uk>"
 
-import cPickle as pickle
 import logging
 import numpy as np
 
@@ -18,7 +17,8 @@ logger = logging.getLogger("explosives")
 class CannonModel(model.BaseModel):
 
     _trained_attributes = ("_coefficients", "_scatter", "_offsets")
-    _data_attributes = ("_wavelengths", "_fluxes", "_flux_uncertainties")
+    _data_attributes = ("_labels", "_wavelengths", "_fluxes",
+        "_flux_uncertainties")
 
     def __init__(self, labels, wavelengths, fluxes, flux_uncertainties,
         verify=True):

### 2015-08-27 explosives/model.py

@@ -5,6 +5,7 @@
 
 __author__ = "Andy Casey <arc@ast.cam.ac.uk>"
 
+import cPickle as pickle
 import numpy as np
 
 def requires_training_wheels(f):
@@ -18,7 +19,7 @@ def requires_training_wheels(f):
     return wrapper
 
 
-def _short_hash(self, contents):
+def _short_hash(contents):
     """ Return a short hash of some iterable contents. """
     return "".join([str(hash(str(item)))[:10] for item in contents])
 
@@ -26,7 +27,8 @@ def _short_hash(self, contents):
 class BaseModel(object):
 
     _trained_attributes = None
-    _data_attributes = None
+    _data_attributes = ("_labels", "_fluxes", "_flux_uncertainties",
+        "_wavelengths")
     
     def __init__(self, labels, fluxes, flux_uncertainties, wavelengths=None,
         verify=True):
@@ -158,7 +160,7 @@ class BaseModel(object):
 
     @property
     def _trained_hash(self):
-        """ Return a hash of the trained state. """
+        """ Return a short joint hash of the trained attributes. """
 
         if not self._trained: return None
         return _short_hash([getattr(self, _) for _ in self._trained_attributes])
@@ -166,15 +168,16 @@ class BaseModel(object):
 
     @property
     def _data_hash(self):
-        """ Return a hash of the data. """
+        """ Return a short joint hash of the data attributes. """
+
         return _short_hash([getattr(self, _) for _ in self._data_attributes])
 
 
-    @model.requires_training_wheels
+    @requires_training_wheels
     def save(self, filename, with_data=False, overwrite=False):
         """
-        Save the (trained) model to disk. This will save the label vector
-        description, the optimised coefficients and scatter, and pivot offsets.
+        Save the (trained) model to disk. This will save all of the relevant
+        training attributes, and optionally, the data attributes.
 
         :param filename:
             The file path where to save the model to.
@@ -183,8 +186,7 @@ class BaseModel(object):
             str
 
         :param with_data: [optional]
-            Save the wavelengths, fluxes and flux uncertainties used to train
-            the model.
+            Also the the data used to train the model.
 
         :type with_data:
             bool
@@ -202,20 +204,23 @@ class BaseModel(object):
             If the model has not been trained, since there is nothing to save.
         """
 
-        assert None not in (self._data_attributes, self._trained_attributes)
-
         contents = [getattr(self, _) for _ in self._trained_attributes]
         contents += [self._data_hash]
         if with_data:
[…]

### 2015-08-27 explosives/utils.py
Progressbar links in with logger more naturally now

@@ -5,7 +5,10 @@
 
 __author__ = "Andy Casey <arc@ast.cam.ac.uk>"
 
+import logging
 import sys
+
+logger = logging.getLogger("explosives")
 
 def progressbar(iterable, message=None, size=100):
     """
@@ -44,7 +47,7 @@ def progressbar(iterable, message=None, size=100):
 
     # Initialise
     if size > 0:
-        sys.stdout.write("\r{}\n".format(message or ""))
+        logger.info((message or "").rstrip())
         sys.stdout.flush()
 
     for i, item in enumerate(iterable):

### 2015-08-27 explosives/explosives.py
Better save/load behaviour (after refactoring), and a from_filename function introduced

@@ -5,7 +5,6 @@
 
 __author__ = "Andy Casey <arc@ast.cam.ac.uk>"
 
-import cPickle as pickle
 import logging
 import numpy as np
 
@@ -18,7 +17,8 @@ logger = logging.getLogger("explosives")
 class ExplosivesModel(cannon.CannonModel):
 
     _save_attributes = ("_coefficients", "_scatter", "_offsets", "_ew_model")
-    _data_attributes = ("_wavelengths", "_fluxes", "_flux_uncertainties")
+    _data_attributes = ("_labels", "_wavelengths", "_fluxes",
+        "_flux_uncertainties")
 
     def __init__(self, labels, wavelengths, fluxes, flux_uncertainties,
         verify=True):
@@ -137,13 +137,21 @@ class ExplosivesModel(cannon.CannonModel):
 
             N_species = len(atomic_lines)
             N_transitions = sum(map(len, atomic_lines.values()))
+
+            msg = []
+            for k, v in atomic_lines.items():
+                msg.append("{0} (species {1}; {2} lines)".format(
+                    k, ", ".join(map(str, set(v["species"]))), len(v)))
+
             logger.info("Including {0} weak lines from {1} elements: {2}".format(
-                N_transitions, N_species, ", ".join(atomic_lines.keys())))
+                N_transitions, N_species, ", ".join(msg)))
 
             # Build the log(X)->EW models (or vice-versa, sigh)
 
             # Estimate the FWHM kernel for each star, or estimate from all stars
             # (We need the FWHM to link the EW to an actual flux value.)
+            # [TODO]
+
             p_sigma = 0.35
 
             # We should calculate the expected EWs (and therefore fluxes) for
@@ -225,7 +233,7 @@ class ExplosivesModel(cannon.CannonModel):
         self._trained = True
         self._coefficients, self._scatter = coefficients, scatter
         self._offsets, self._ew_model = offsets, ew_model
-        
+
         return (coefficients, scatter, offsets, ew_model)
 
 

### 2015-08-27 explosives/model.py

@@ -7,6 +7,8 @@ __author__ = "Andy Casey <arc@ast.cam.ac.uk>"
 
 import cPickle as pickle
 import numpy as np
+import os
+
 
 def requires_training_wheels(f):
     """
@@ -212,15 +214,14 @@ class BaseModel(object):
         if os.path.exists(filename) and not overwrite:
             raise IOError("filename '{}' exists, asked not to overwrite".format(
                 filename))
-        
+
         with open(filename, "w") as fp:
             pickle.dump(contents, fp, -1)
 
         return True
 
-
-    @classmethod
-    def load(cls, filename, verify=True):
+    
+    def load(self, filename, verify=True):
         """
         Load a trained model from disk.
 
@@ -253,26 +254,83 @@ class BaseModel(object):
             contents = pickle.load(fp)
 
         # Contents is: trained attributes, data hash, [data trained on]
-        trained_contents = dict(zip(cls._trained_attributes, contents))
+        trained_contents = dict(zip(self._trained_attributes, contents))
         N = len(trained_contents)
         expected_data_hash = contents[N]
 
-        _repr_data_attributes = [_[1:] for _ in cls._data_attributes]
         if len(contents) > N + 1:
             # There was data as well.
-            data_contents = dict(zip(_repr_data_attributes, contents[N + 1:]))
+            data_contents = dict(zip(self._data_attributes, contents[N + 1:]))
             if verify and expected_data_hash is not None:
                 actual_data_hash = _short_hash(data_contents)
                 if actual_data_hash != expected_data_hash:
                     raise ValueError("expected data hash ({0}) is different "\
                         "({1})".format(expected_data_hash, actual_data_hash))
 
-        else:
-            data_contents = dict(zip(_repr_data_attributes,
-                [getattr(self, k) for k in cls._data_attributes]))
+            # Set the data attributes.
+            for k, v in data_contents.items():
+                setattr(self, k, v)
 
-        # Initialise a model.
-        model = cls(**data_contents)
+        # Set the training attributes.
+        for k, v in trained_contents.items():
+            setattr(self, k, v)
+
+        self._trained = True
+        return True
+
+
+    @classmethod
+    def from_filename(cls, filename, verify=True):
+        """
+        Initialise a trained model from disk. The saved model must include data.
+
+        :param filename:
+            The file path where to load the model from.
+
+        :type filename:
+            str
+
+        :param verify: [optional]
+            Verify whether the hashes in the stored filename match what is
[…]

### 2015-08-28 explosives/cannon.py
IT'S ALIVE

@@ -254,6 +254,18 @@ class CannonModel(model.BaseModel):
 
     @property
     @model.requires_training_wheels
+    def labels(self):
+        """
+        Return a list of the labels involved in this model.
+        """
+
+        _, labels = self._get_linear_indices(self._label_vector_description,
+            full_output=True)
+        return list(labels)
+
+
+    @property
+    @model.requires_training_wheels
     def label_residuals(self):
         """
         Calculate the residuals of the inferred labels using the trained model.
@@ -274,21 +286,20 @@ class CannonModel(model.BaseModel):
         except AttributeError:
             None
 
-        label_indices, label_names = self._get_linear_indices(
-            self._label_vector_description, full_output=True)
-        expected_labels = np.zeros((self._fluxes.shape[0], len(label_names)))
-        inferred_labels = np.zeros((self._fluxes.shape[0], len(label_names)))
+        labels = self.labels
+        expected_labels = np.zeros((self._fluxes.shape[0], len(labels)))
+        inferred_labels = np.zeros((self._fluxes.shape[0], len(labels)))
 
         for i, (flux, uncertainty) \
         in enumerate(zip(self._fluxes, self._flux_uncertainties)):
             inferred = self.solve_labels(flux, uncertainty)
-            for j, label_name in enumerate(label_names):
+            for j, label_name in enumerate(labels):
                 expected_labels[i, j] = self._labels[label_name][i]
                 inferred_labels[i, j] = inferred[label_name]
 
         # Cache for future, unless the training state changes.
         self._residuals_hash = self._trained_hash
-        self._residuals_cache = (label_names, expected_labels, inferred_labels)
+        self._residuals_cache = (labels, expected_labels, inferred_labels)
         return self._residuals_cache
 
 
@@ -386,8 +397,8 @@ class CannonModel(model.BaseModel):
             if len(term) == 1 and term[0][1] == 1:
                 names.append(term[0][0])
                 indices.append(i)
-        names, indices = tuple(names), np.array(indices)
-
+        indices = np.array(indices)
+        
         if full_output:
             return (indices, names)
         return indices

### 2015-08-28 explosives/explosives.py

@@ -7,6 +7,7 @@ __author__ = "Andy Casey <arc@ast.cam.ac.uk>"
 
 import logging
 import numpy as np
+from collections import OrderedDict
 
 import scipy.optimize as op
 from . import (atomic, cannon, model, plot, utils)
@@ -16,7 +17,8 @@ logger = logging.getLogger("explosives")
 
 class ExplosivesModel(cannon.CannonModel):
 
-    _save_attributes = ("_coefficients", "_scatter", "_offsets", "_ew_model")
+    _trained_attributes = ("_coefficients", "_scatter", "_offsets", "_ew_model",
+        "_weak_line_mask", "_label_vector_description", "_atomic_lines")
     _data_attributes = ("_labels", "_wavelengths", "_fluxes",
         "_flux_uncertainties")
 
@@ -55,6 +57,22 @@ class ExplosivesModel(cannon.CannonModel):
 
         super(self.__class__, self).__init__(labels, wavelengths, fluxes,
             flux_uncertainties, verify)
+
+
+    @property
+    @model.requires_training_wheels
+    def labels(self):
+        """
+        Return a list of the labels involved in this model. This includes any
+        labels in the description of the label vector, as well as any individual
+        abundance labels.
+        """
+
+        _, labels = self._get_linear_indices(self._label_vector_description,
+            full_output=True)
+        if self._atomic_lines is not None:
+            labels += self._atomic_lines.keys()
+        return labels
 
 
     def train(self, label_vector_description, atomic_lines=None, X_H=False,
@@ -111,8 +129,9 @@ class ExplosivesModel(cannon.CannonModel):
             in each pixel, and the label offsets.
         """
 
-        # We'll need this later; store it.
-        self._atomic_lines = atomic_lines
+        # We'll need this later; store it. First sort by the species number.
+        self._atomic_lines = OrderedDict(sorted(atomic_lines.items(), 
+            key=lambda _: min(_[1]["species"])))
         self._label_vector_description = label_vector_description
 
         # Build the label vector array. Note that if `atomic_lines` are given,
@@ -133,13 +152,13 @@ class ExplosivesModel(cannon.CannonModel):
         weak_line_fluxes = np.ones((N_stars, N_pixels))
         
         # Any atomic lines to model?
-        if _check_atomic_lines(self._labels, atomic_lines):
+        if _check_atomic_lines(self._labels, self._atomic_lines):
 
-            N_species = len(atomic_lines)
[…]

### 2015-08-28 explosives/model.py

@@ -9,7 +9,6 @@ import cPickle as pickle
 import numpy as np
 import os
 
-
 def requires_training_wheels(f):
     """
     A decorator for model functions where the model needs training first.
@@ -158,6 +157,12 @@ class BaseModel(object):
                         " label '{1}' - to ignore this use verify=False".format(
                             character, column))
         return True
+
+
+    @property
+    def labels(self):
+        """ This should be overwritten by the parent model. """
+        raise WTFError("is this your first day?")
 
 
     @property
@@ -338,3 +343,5 @@ class BaseModel(object):
 
         model._trained = True
         return model
+
+

### 2015-08-28 sandbox.py

@@ -1,0 +1,100 @@
+#!/usr/bin/env python
+# -*- coding: utf-8 -*-
+
+""" Explosives sandbox. """
+
+__author__ = "Andy Casey <arc@ast.cam.ac.uk>"
+
+import numpy as np
+from astropy.table import Table
+from explosives import explosives
+
+DATA_PREFIX = "/Users/arc/research/luminosity-cannon/data/APOGEE-Hipparcos"
+stars = Table.read("{}.fits.gz".format(DATA_PREFIX))
+
+
+import scipy.optimize as op
+def vac2air(wave,sdssweb=False):
+    """
+    NAME:
+       vac2air
+    PURPOSE:
+       Convert from vacuum to air wavelengths (See Allende Prieto technical note: http://hebe.as.utexas.edu/apogee/docs/air_vacuum.pdf)
+    INPUT:
+       wave - vacuum wavelength in \AA
+       sdssweb= (False) if True, use the expression from the SDSS website (http://classic.sdss.org/dr7/products/spectra/vacwavelength.html)
+    OUTPUT:
+       air wavelength in \AA
+    HISTORY:
+       2014-12-04 - Written - Bovy (IAS)
+       2015-04-27 - Updated to CAP note expression - Bovy (IAS)
+    """
+    if sdssweb:
+        return wave/(1.+2.735182*10.**-4.+131.4182/wave**2.+2.76249*10.**8./wave**4.)
+    else:
+        return wave/(1.+0.05792105/(238.0185-(10000./wave)**2.)+0.00167917/(57.362-(10000./wave)**2.))
+
+def air2vac(wave,sdssweb=False):
+    """
+    NAME:
+       air2vac
+    PURPOSE:
+       Convert from air to vacuum wavelengths (See Allende Prieto technical note: http://hebe.as.utexas.edu/apogee/docs/air_vacuum.pdf)
+    INPUT:
+       wave - air wavelength in \AA
+       sdssweb= (False) if True, use the expression from the SDSS website (http://classic.sdss.org/dr7/products/spectra/vacwavelength.html)
+    OUTPUT:
+       vacuum wavelength in \AA
+    HISTORY:
+       2014-12-04 - Written - Bovy (IAS)
+       2015-04-27 - Updated to CAP note expression - Bovy (IAS)
+    """
[…]

### 2015-08-28 fireworks/__init__.py
BOOM #1

@@ -10,11 +10,8 @@ import logging
 from numpy import RankWarning
 from warnings import simplefilter
 
-logger = logging.getLogger("explosives")
+logger = logging.getLogger("fireworks")
 logger.setLevel(logging.INFO)
-
-#logging.basicConfig(level=logging.INFO, 
-#    format="%(asctime)s [%(levelname)s] %(message)s")
 
 simplefilter("ignore", RankWarning)
 simplefilter("ignore", RuntimeWarning)

### 2015-08-28 fireworks/atomic.py

@@ -11,7 +11,7 @@ import scipy.optimize as op
 
 from oracle import photospheres, synthesis
 
-logger = logging.getLogger("explosives")
+logger = logging.getLogger("fireworks")
 
 
 def approximate_atomic_transitions(stellar_parameters, transitions, X_H=False,

### 2015-08-28 fireworks/cannon.py

@@ -11,7 +11,7 @@ import numpy as np
 import scipy.optimize as op
 from . import (model, plot, utils)
 
-logger = logging.getLogger("explosives")
+logger = logging.getLogger("fireworks")
 
 
 class CannonModel(model.BaseModel):

### 2015-08-28 fireworks/fireworks.py

@@ -12,10 +12,10 @@ from collections import OrderedDict
 import scipy.optimize as op
 from . import (atomic, cannon, model, plot, utils)
 
-logger = logging.getLogger("explosives")
+logger = logging.getLogger("fireworks")
 
 
-class ExplosivesModel(cannon.CannonModel):
+class FireworksModel(cannon.CannonModel):
 
     _trained_attributes = ("_coefficients", "_scatter", "_offsets", "_ew_model",
         "_weak_line_mask", "_label_vector_description", "_atomic_lines")

### 2015-08-28 fireworks/utils.py

@@ -8,7 +8,7 @@ __author__ = "Andy Casey <arc@ast.cam.ac.uk>"
 import logging
 import sys
 
-logger = logging.getLogger("explosives")
+logger = logging.getLogger("fireworks")
 
 def progressbar(iterable, message=None, size=100):
     """

### 2015-08-28 sandbox.py

@@ -1,13 +1,13 @@
 #!/usr/bin/env python
 # -*- coding: utf-8 -*-
 
-""" Explosives sandbox. """
+""" fireworks sandbox. """
 
 __author__ = "Andy Casey <arc@ast.cam.ac.uk>"
 
 import numpy as np
 from astropy.table import Table
-from explosives import explosives
+from fireworks import fireworks
 
 DATA_PREFIX = "/Users/arc/research/luminosity-cannon/data/APOGEE-Hipparcos"
 stars = Table.read("{}.fits.gz".format(DATA_PREFIX))
@@ -73,7 +73,7 @@ stellar_parameters = np.vstack([stars["TEFF"], stars["LOGG"],
 #atomic.approximate_atomic_transitions(stellar_parameters, transitions)
 
 
-# Create ExplosivesModel
+# Create FireworksModel
 fluxes = np.memmap("{}-flux.memmap".format(DATA_PREFIX), mode="r", dtype=float)
 flux_uncertainties = np.memmap("{}-flux-uncertainties.memmap".format(DATA_PREFIX),
     mode="r", dtype=float)
@@ -89,12 +89,12 @@ atomic_lines = {
 label_vector_description = "TEFF^4 TEFF^3 TEFF^2 TEFF LOGG LOGG^2 TEFF*LOGG TEFF^2*LOGG TEFF*LOGG^2 PARAM_M_H PARAM_M_H*TEFF PARAM_M_H*TEFF^2 PARAM_ALPHA_M PARAM_M_H*PARAM_ALPHA_M" 
 
 if False:
-    m = explosives.ExplosivesModel(stars, wavelengths, fluxes, flux_uncertainties)
+    m = fireworks.FireworksModel(stars, wavelengths, fluxes, flux_uncertainties)
     stuff = m.train(label_vector_description, atomic_lines, X_H=True)
     m.save("temp.pkl", with_data=True, overwrite=True)
 
 else:
-    m = explosives.ExplosivesModel.from_filename("temp.pkl")
+    m = fireworks.FireworksModel.from_filename("temp.pkl")
 
 label_names, expected_labels, inferred_labels = m.label_residuals
 raise a

### 2015-08-28 fireworks/atomic.py
stop being shit

@@ -66,23 +66,10 @@ def approximate_atomic_transitions(stellar_parameters, transitions, X_H=False,
         dict
     """
 
-    import cPickle as pickle
-
-    if False:
-        # Calculate abundances for all lines in all stars.
-        ews, abundances = _atomic_line_abundances(stellar_parameters, transitions,
-            X_H=X_H, ew_points=ew_points, photosphere_kwargs=photosphere_kwargs,
-            synthesis_kwargs=synthesis_kwargs, **kwargs)
-
-        with open("tmp.pkl", "wb") as fp:
-            pickle.dump((ews, abundances), fp, -1)
-    else:
-        with open("tmp.pkl", "rb") as fp:
-            ews, abundances = pickle.load(fp)
-
-        if X_H:
-            logger.warn("ASSUMING ALL FE")
-            abundances -= 7.5
+    # Calculate abundances for all lines in all stars.
+    ews, abundances = _atomic_line_abundances(stellar_parameters, transitions,
+        X_H=X_H, ew_points=ew_points, photosphere_kwargs=photosphere_kwargs,
+        synthesis_kwargs=synthesis_kwargs, **kwargs)
 
     # f(teff, logg, EW) --> logX
     full_output = kwargs.pop("full_output", False)

### 2015-08-28 fireworks/fireworks.py
Predict spectra even with weak atomic lines present in the model

@@ -266,9 +266,10 @@ class FireworksModel(cannon.CannonModel):
 
         :param labels:
             The labels required for the trained model. This should be a N-length
-            list matching the number of unique terms in the model, in the order
-            given by the `self._get_linear_indices` function. Alternatively,
-            labels can be explicitly given as keyword arguments.
+            list matching the number of unique terms in the model, including any
+            atomic (weak) line abundances in the order given by the `self.labels`
+            property. Alternatively, labels can be explicitly given as keyword
+            arguments.
 
         :type labels:
             list
@@ -279,31 +280,64 @@ class FireworksModel(cannon.CannonModel):
         :raises TypeError:
             If the model is not trained.
         """
-        raise NotImplementedError
         
         try:
             labels[0]
         except (TypeError, IndexError):
             labels = [labels]
 
-        indices, names = self._get_linear_indices(
-            self._label_vector_description, full_output=True)
+        label_names = self.labels
         if labels is None:
             # Must be given as keyword arguments.
-            labels = [labels_as_kwargs[name] for name in names]
+            labels = [labels_as_kwargs[name] for name in label_names]
 
         else:
-            if len(labels) != len(names):
+            if len(labels) != len(label_names):
                 raise ValueError("expected number of labels is {0}, and {1} "
-                    "were given: {2}".format(len(names), len(labels),
-                        ", ".join(names)))
+                    "were given: {2}".format(len(label_names), len(labels),
+                        ", ".join(label_names)))
 
+        # Generate the Cannon-ical flux.
         label_vector_indices = self._parse_label_vector_description(
             self._label_vector_description, return_indices=True,
-            __columns=names)
+            __columns=label_names)
 
-        return np.dot(self._coefficients, cannon._build_label_vector_rows(
+        fluxes = np.dot(self._coefficients, cannon._build_label_vector_rows(
             label_vector_indices, labels).T).flatten()
+
+        # Include treatment of any atomic lines.
+        if self._atomic_lines is not None:
+
+            N = len(self._atomic_lines)
+            stellar_parameters = labels[:3]
+            logger.warn("No explicit reference to stellar parameters")
+            # [TODO] No explicit reference to stellar parameter labels  
+
+            weak_line_fluxes = np.ones(cannonical_flux.size)
+            for i, (label, abundance) \
[…]

### 2015-08-28 sandbox.py
Trying atomic abundances with many elements

@@ -74,6 +74,7 @@ stellar_parameters = np.vstack([stars["TEFF"], stars["LOGG"],
 
 
 # Create FireworksModel
+"""
 fluxes = np.memmap("{}-flux.memmap".format(DATA_PREFIX), mode="r", dtype=float)
 flux_uncertainties = np.memmap("{}-flux-uncertainties.memmap".format(DATA_PREFIX),
     mode="r", dtype=float)
@@ -97,4 +98,113 @@ else:
     m = fireworks.FireworksModel.from_filename("temp.pkl")
 
 label_names, expected_labels, inferred_labels = m.label_residuals
+"""
+
+
+# Create a FireworksModel with many elements.
+fluxes = np.memmap("{}-flux.memmap".format(DATA_PREFIX), mode="r", dtype=float)
+flux_uncertainties = np.memmap("{}-flux-uncertainties.memmap".format(DATA_PREFIX),
+    mode="r", dtype=float)
+wavelengths = np.memmap("{}-wavelength.memmap".format(DATA_PREFIX), mode="r",
+    dtype=float)
+
+fluxes = fluxes.reshape((len(stars), -1))
+flux_uncertainties = flux_uncertainties.reshape(fluxes.shape)
+
+# Let's try with 5 elements from different nucleosynthetic pathways, all with
+# different number of lines and different line qualities.
+
+# Mg 1, Al 1, Si I, Cr I, Co I
+
+# From Smith et al. (2013)
+MgI_transitions = Table(np.array([
+    [15740.716, 12.0,   5.931, -0.262],
+    [15748.9,   12.0,   5.932, +0.276],
+    [15765.8,   12.0,   5.933, +0.504],
+    [15879.5,   12.0,   5.946, -1.248],
+    [15886.2,   12.0,   5.946, -1.555],
+    [15889.485, 12.0,   5.946, -2.013],
+    [15954.477, 12.0,   6.588, -0.807]
+    ]), names=["wavelength", "species", "excitation_potential", "loggf"])
+MgI_transitions["wavelength"] = np.array(map(air2vac, MgI_transitions["wavelength"]))
+
+AlI_transitions = Table(np.array([
+    [16718.957, 13.0,   4.085, +0.290],
+    [16763.359, 13.0,   4.087, -0.524],
+    ]), names=["wavelength", "species", "excitation_potential", "loggf"])
[…]

### 2015-08-28 article/ms.tex
Starting to describe The Cannon, and contrast to current methods

@@
 
+% Convenient colloquiualisms.
 \newcommand\tc{\textit{The Cannon}}
-\title[Cannon Chemistry]{Extending \tc{} to Individual Chemical Abundances} 
+
+% Convenient maths.
+\newcommand\lv{\mathbf{l_n}}
+\newcommand\cv{{\mathbf \theta}_\lambda}
+\newcommand\given{|}
+
+\title[Cannon Chemistry]{Detailed Chemical Abundances with \tc} 
 % Some arrangement of the following authors:
-\author[Casey et al.]{A.~R.~Casey$^1$, M.~Ness$^2$, G.~Gilmore$^1$, D.~W.~Hogg$^{2,3,4}$, H.~W.~Rix$^2$ \\ 
+\author[Casey et al.]{A.~R.~Casey$^1$, M.~Ness$^2$, G.~Gilmore$^1$,
+    D.~W.~Hogg$^{2,3,4}$, H.~W.~Rix$^2$ \\ 
 $^1$Institute of Astronomy, University of Cambridge, Madingley Road, Cambridge
@@
 \begin{abstract}
+% large surveys spectroscopically planned.
+% poses computing challenge.
+% commonalityies between them: all R ~ 20,000, trying to get abundances from
+% weak lines, which are linearly responsive (therefore best indicators) of
+% abundance.
+% TC provides a purely data-driven approach to stellar parameter (label)
+% determination.
+% fast, accurate, limited by systematics (which are low), meaning it works well
+% in low S/N. Perfect for survey work.
+% Simply extending dimensionality of the Cannon to include individual abundances
+% can create enormous difficulties, both in potentially overfitting and curse of DIm.
+% These extensions also imply a purely data-driven approach to abundances.
+% Here we include radiative transfer physics to explain why a polynomial extension
+% works so well, and provide a simple model to relate individual chemical
+% abundance (and teh stellar parameter labels) to a flux value.
+% We provide a direct link between model atmosphere physics and data-driven
+% approaches, allowing the data to inform models.
+% We demonstrate that individual abundances can be determined with a precision of
+% X using our models. These determinations are fast: we estimate all 15 APOGEE 
+% elemental abundances (and stellar parameters) for 60,000 stars on a modern
+% laptop in 5 minutes.
+
 \end{abstract}
@@
 
+% it's awesome: fast, accurate and precise (given the models published thus far)
+% impressively it does all this whilst ignoring all advances made in stellar physics over the past 100 years.
+
+\section{Model}
+
+\tc{} provides a purely data-driven approach to stellar label determination. The
+spectral model describes the flux at every pixel $f_{n\lambda}$ to be a function
+of a coefficient vector $\cv$, and a label vector $\lv$:
+
+\begin{equation}
+    f_{n\lambda} = g(\lv\given\cv) + {\rm noise}.
+\end{equation}
+
+This form is flexible: the label vector $\lv$ may be linear in just a few labels
+or a complex combination of all $K$ available labels. The underlying principle
+is that the flux for any pixel from any star is predictable with a very simple
+model (generally a polynomial), given the labels. Indeed, \citet{Ness2015}
+employ a quadratic-in-labels (see their Equation 8) spectral model using stellar
+parameters (effective temperature $T_{\rm eff}$, surface gravity $\log{g}$, and 
+metallicity [M/H]) and demonstrate by cross-validation that these labels can be 
+measured with a precision of 95~K in $T_{\rm eff}$, 0.24 in $\log{g}$, 0.08 in 
+[M/H], with negligible biases.
+
+Utilising a simple polynomial model to predict spectral fluxes is in stark
+contrast to parallel efforts in radiative transfer of realistic (3D) stellar 
+photospheres. Calculations of this nature are extremely expensive. Indeed, even
+with simplications of these models (e.g., interpolated 1D photospheres),
+calculating integrated flux requires knowledge of species opacities, detailed 
+atomic and molecular information for all contributing elements, solving for
+the equation of state, and performing radiative transfer calculations. Given the
+simplicity of \tc{} and a quadratic-in-labels spectral model, it is remarkable
+that these complex calculations can be simplified so much, whilst maintaining
+the typical fidelity reported for high-resolution spectroscopic studies.
+
+% existing simple model. replaces expensive synthesis, involving model atmosphere
+% generation, t-tau relations, opacities, equation of state, integrated line-of-sight
+% and radiative transfer, etc.
+
+% before extending this model to include individual abundances, it is worthwhile
+% demonstrating (from a physics perspective) *why* the cannon works so well.
+
+% APOGEE, R of 20,000, 
+% LAMOST, higher Rs
+% super high-resolution also works.
+
+\section{Analysis}
+% Detailed chemical abundances with APOGEE.
+
 \section{Conclusions}
@@
 \section*{Acknowledgments}
-This research was fostered by discussions at the \textit{New Milky Way} conference hosted by the Munich Institute for Astro- and Particle Physics, and the \textit{Stellar Streams in the Local Universe} conference hosted at Ringberg Castle by the Max-Planck-Institut F\"ur Astronomie.
-This research made use of Astropy, a community-developed core Python package for Astronomy \citep{astropy}.
+This research was fostered by discussions at the \textit{New Milky Way}
+conference hosted by the Munich Institute for Astro- and Particle Physics, and
+the \textit{Stellar Streams in the Local Universe} conference hosted at Ringberg
+Castle by the Max-Planck-Institut F\"ur Astronomie. This research made use of 
+Astropy, a community-developed core Python package for Astronomy \citep{astropy}.
 

### 2015-08-28 article/ms.tex
Some text cleanup and boldmath all the things [ci skip]

@@
 
+\usepackage{amsmath,amsbsy}
+
 % Convenient colloquiualisms.
+\newcommand\article{\textit Article}
 \newcommand\tc{\textit{The Cannon}}
@@
 % Convenient maths.
-\newcommand\lv{\mathbf{l_n}}
-\newcommand\cv{{\mathbf \theta}_\lambda}
+\newcommand\K{$\mathbf{K}$}
+\newcommand\lv{\bf{\boldsymbol\ell_n}}
+\newcommand\cv{{\boldsymbol\theta}_\lambda}
 \newcommand\given{|}
@@
 parameters (effective temperature $T_{\rm eff}$, surface gravity $\log{g}$, and 
-metallicity [M/H]) and demonstrate by cross-validation that these labels can be 
+metallicity [M/H]\footnote{\citet{Ness2015} use [Fe/H] in $\lv$, but for the 
+sake of distinguishing between overall metallicity and individual abundances,
+we opt for [M/H] to describe overall metallicity throughout this \article{}})
+and demonstrate by cross-validation that these labels can be 
 measured with a precision of 95~K in $T_{\rm eff}$, 0.24 in $\log{g}$, 0.08 in 
-[M/H], with negligible biases.
+[M/H], with negligible biases. It is important to note that these uncertainties
+are comparable to those reported by \textit{curve-of-growth} techniques, which
+rely on a model atmosphere and the changing strength of atomic absorption lines.
 
-Utilising a simple polynomial model to predict spectral fluxes is in stark
-contrast to parallel efforts in radiative transfer of realistic (3D) stellar 
+Utilising a simple polynomial model to predict spectral fluxes is a distinctive
+method from parallel efforts in radiative transfer of realistic (3D) stellar 
 photospheres. Calculations of this nature are extremely expensive. Indeed, even
@@
 \begin{thebibliography}{99}
-%\bibitem[\protect\citeauthoryear{Baird}{1981}]{b1} Baird S.R., 1981,
-%ApJ, 245, 208
 \bibitem[Astropy Collaboration et al.(2013)]{astropy} Astropy Collaboration, Robitaille, T.~P., Tollerud, E.~J., et al.\ 2013, \aap, 558, AA33
+
+\bibitem[Ness et al.(2015)]{Ness2015} Ness, M., Hogg, D.~W., 
+Rix, H.-W., Ho, A.~Y.~Q., \& Zasowski, G.\ 2015, \apj, 808, 16 
 \end{thebibliography}

### 2015-08-28 article/ms.tex
Good enough for government work

@@
 \newcommand\K{$\mathbf{K}$}
-\newcommand\lv{\bf{\boldsymbol\ell_n}}
+\newcommand\lv{\mathbf{\boldsymbol\ell_n}}
 \newcommand\cv{{\boldsymbol\theta}_\lambda}
@@
 [M/H], with negligible biases. It is important to note that these uncertainties
-are comparable to those reported by \textit{curve-of-growth} techniques, which
-rely on a model atmosphere and the changing strength of atomic absorption lines.
+are comparable to those reported from using \textit{curve-of-growth} on similar 
+data sets (e.g., \citet{X,Y}).
 
@@
 the typical fidelity reported for high-resolution spectroscopic studies.
+
+Before extending this framework to include individual chemical abundances, it is
+a useful exercise to understand \textit{why} a simple polynomial model can 
+approximate spectral synthesis calculations with comparable fidelity. The reason
+extends past model flexibility: one could imagine a similar coefficient vector
+$\cv$ with a different functional form of the label vector $\lv$ (e.g., an
+extreme case might be $\lv \equiv [1, {\rm [M/H]}^{T_{\rm eff}},
+\log{g}^{T_{\rm eff}}]$) which would probably be incapable of describing
+spectral fluxes with any fidelity.
+
+% Describe the model of EW formation
+
+% How this changes as a function of teff, logg, feh
+
+% dealing with integrated EW.
+% APOGEE is R ~ 20,000
 

### 2015-08-28 sandbox.py

@@ -147,6 +147,7 @@ SiI_transitions = Table(np.array([
 SiI_transitions["wavelength"] = np.array(map(air2vac, SiI_transitions["wavelength"]))
 
 
+# It seems this is not in APOGEE.
 CrI_transitions = Table(np.array([
     [15680.063, 24.0,   4.697, +0.179],
     [15860.214, 24.0,   4.697, -0.012]
@@ -159,23 +160,43 @@ CoI_transitions = Table(np.array([
     ]), names=["wavelength", "species", "excitation_potential", "loggf"])
 CoI_transitions["wavelength"] = np.array(map(air2vac, CoI_transitions["wavelength"]))
 
+VI_transitions = Table(np.array([
+    [15924., 23.0, 2.138, -1.175]
+    ]), names=["wavelength", "species", "excitation_potential", "loggf"])
+VI_transitions["wavelength"] = np.array(map(air2vac, VI_transitions["wavelength"]))
 
 
+
+Ni1_transitions = Table(np.array([
+    [15605.680, 28.0, 5.299, -0.376],
+    [15632.654, 28.0, 5.305, -0.106],
+    [16584.439, 28.0, 5.299, -0.528],
+    [16589.295, 28.0, 5.469, -0.600],
+    [16673.711, 28.0, 6.029, +0.317],
+    [16815.471, 28.0, 5.305, -0.606],
+    [16818.760, 28.0, 6.039, +0.311],
+    ]), names=["wavelength", "species", "excitation_potential", "loggf"])
 
 atomic_lines = {
     "MG_H": MgI_transitions,
-    "AL_H": AlI_transitions
+    "AL_H": AlI_transitions,
+    "SI_H": SiI_transitions,
+#    "CR_H": CrI_transitions, # Not in APOGEE
+    "V_H": VI_transitions,
+#    "CO_H": CoI_transitions, # Not in APOGEE
+    "NI_H": Ni1_transitions,
+
 }
 label_vector_description = "TEFF^4 TEFF^3 TEFF^2 TEFF LOGG LOGG^2 TEFF*LOGG TEFF^2*LOGG TEFF*LOGG^2 PARAM_M_H PARAM_M_H*TEFF PARAM_M_H*TEFF^2 PARAM_ALPHA_M PARAM_M_H*PARAM_ALPHA_M" 
 
 
-temp_filename = "fireworks-Mg+Al.pkl"
[…]

### 2015-08-28 article/ms.tex
First draft of abstract

@@
 \begin{abstract}
-% large surveys spectroscopically planned.
-% poses computing challenge.
-% commonalityies between them: all R ~ 20,000, trying to get abundances from
-% weak lines, which are linearly responsive (therefore best indicators) of
-% abundance.
-% TC provides a purely data-driven approach to stellar parameter (label)
-% determination.
-% fast, accurate, limited by systematics (which are low), meaning it works well
-% in low S/N. Perfect for survey work.
-% Simply extending dimensionality of the Cannon to include individual abundances
-% can create enormous difficulties, both in potentially overfitting and curse of DIm.
-% These extensions also imply a purely data-driven approach to abundances.
-% Here we include radiative transfer physics to explain why a polynomial extension
-% works so well, and provide a simple model to relate individual chemical
-% abundance (and teh stellar parameter labels) to a flux value.
-% We provide a direct link between model atmosphere physics and data-driven
-% approaches, allowing the data to inform models.
-% We demonstrate that individual abundances can be determined with a precision of
-% X using our models. These determinations are fast: we estimate all 15 APOGEE 
-% elemental abundances (and stellar parameters) for 60,000 stars on a modern
-% laptop in 5 minutes.
-
+More than 15 million spectra are expected to be acquired by stellar surveys in
+the coming decade. The detailed chemical abundance analysis of these data poses
+a significant computational challenge. \tc provides a purely data-driven
+approach to stellar parameter (label) determination. Cross-validation tests
+demonstrate that the precision achievable is comparable to existing methods that
+rely on model atmospheres and spectral synthesis, but \tc is $\sim300,000\times$ faster, making it perfectly apt for surveys. Simply extending the label vector
+dimensionality of \tc to include individual abundances can create enormous
+model difficulties, severely limiting the interpretability of the results. Here
+we describe an alternative approach to including individual element abundances
+in \tc. Using modest simplifications of the physics of spectral line formation,
+we demonstrate precisely \textit{why} an approach like \tc works so effectively,
+and provide a physically-motivated model that describes flux values, given the
+stellar parameters. The inclusion of individual chemical abundances does not
+impact on model flexibility, allowing for simultaneous training on detailed
+chemical abundances, as well as arbitrary labels where the impact on the spectra
+is unknown or poorly understood (e.g., mass, age). Crucially, we provide a link
+between data-driven models and parallel works on model atmospheres and spectral
+line formation, allowing for data to inform models. We demonstrate that
+detailed chemical abundances can be measured with a precision of 0.0X-0.0Y~dex
+in Z APOGEE spectra on a modern CPU in just one minute.
 \end{abstract}

### 2015-09-01 article/ms.tex
Formatting

@@
 \newcommand\article{\textit Article}
-\newcommand\tc{\textit{The Cannon}}
+\newcommand\tc{\textit{The Cannon} }
 
@@
 demonstrate that the precision achievable is comparable to existing methods that
-rely on model atmospheres and spectral synthesis, but \tc is $\sim300,000\times$ faster, making it perfectly apt for surveys. Simply extending the label vector
+rely on model atmospheres and spectral synthesis, but \tc is $\sim$300,000 times
+faster, making it perfectly apt for surveys. Simply extending the label vector
 dimensionality of \tc to include individual abundances can create enormous

### 2015-09-07 fireworks/utils.py
Updated progressbar and added label vector utility

@@ -5,10 +5,13 @@
 
 __author__ = "Andy Casey <arc@ast.cam.ac.uk>"
 
-import logging
 import sys
+import logging
+from time import time
+from collections import Counter
+from itertools import combinations_with_replacement
 
-logger = logging.getLogger("fireworks")
+logger = logging.getLogger("cannon")
 
 def progressbar(iterable, message=None, size=100):
     """
@@ -34,15 +37,17 @@ def progressbar(iterable, message=None, size=100):
         int
     """
 
+    t_init = time()
     count = len(iterable)
-    def _update(i):
+    def _update(i, t=None):
         if 0 >= size: return
         increment = max(1, int(count / 100))
-        if i == 0 or i % increment == 0:
-            sys.stdout.write("\r[{done}{not_done}] {percent:3.0f}%".format(
-                done="=" * int((i + 1) / increment),
-                not_done=" " * int((count - i - 1)/increment),
-                percent=100. * (i + 1)/count))
+        if i % increment == 0 or i in (0, count):
+            sys.stdout.write("\r[{done}{not_done}] {percent:3.0f}%{t}".format(
+                done="=" * int(i/increment),
+                not_done=" " * int((count - i)/increment),
+                percent=100. * i/count,
+                t="" if t is None else " ({0:.0f}s)".format(t-t_init)))
             sys.stdout.flush()
 
     # Initialise
@@ -52,8 +57,62 @@ def progressbar(iterable, message=None, size=100):
 
     for i, item in enumerate(iterable):
         yield item
-        _update(i + 1)
+        _update(i)
 
     if size > 0:
+        _update(count, time())
         sys.stdout.write("\r\n")
         sys.stdout.flush()
+
+
+def label_vector(labels, order, cross_term_order=0, mul="*", pow="^"):
+    """
+    Build a label vector description.
+
+    :param labels:
+        The labels to use in describing the label vector.
+
+    :type labels:
+        iterable
+
+    :param order:
+        The maximum order of the terms (e.g., order 3 implies A^3 is a term).
+
+    :type order:
+        int
+
+    :param cross_term_order: [optional]
+        The maximum order of the cross-terms (e.g., cross_term_order 2 implies
+        A^2*B is a term).
+    
+    :type cross_term_order:
+        int
+
+    :param mul: [optional]
+        The operator to use to represent multiplication in the description of 
+        the label vector.
+
+    :type mul:
+        str
+
+    :param pow: [optional]
[…]

### 2015-09-07 fireworks/cannon.py
Parallelise training, include cross-validation codes, and various improvements

@@ -1,14 +1,16 @@
 #!/usr/bin/env python
 # -*- coding: utf-8 -*-
 
-""" Cannon Chemistry """
+""" The Cannon. """
 
 __author__ = "Andy Casey <arc@ast.cam.ac.uk>"
 
 import logging
+import multiprocessing as mp
 import numpy as np
-
 import scipy.optimize as op
+from astropy.table import Table
+
 from . import (model, plot, utils)
 
 logger = logging.getLogger("fireworks")
@@ -16,9 +18,11 @@ logger = logging.getLogger("fireworks")
 
 class CannonModel(model.BaseModel):
 
-    _trained_attributes = ("_coefficients", "_scatter", "_offsets")
-    _data_attributes = ("_labels", "_wavelengths", "_fluxes",
-        "_flux_uncertainties")
+    _trained_attributes \
+        = ("_coefficients", "_scatter", "_offsets", "_label_vector_description")
+    _data_attributes \
+        = ("_labels", "_wavelengths", "_fluxes", "_flux_uncertainties")
+
 
     def __init__(self, labels, wavelengths, fluxes, flux_uncertainties,
         verify=True):
@@ -32,17 +36,17 @@ class CannonModel(model.BaseModel):
             :class:`~astropy.table.Table`
 
         :param wavelengths:
-            The wavelengths of the given pixels.
+            The common wavelength values of the pixels.
 
         :type wavelengths:
-            :class:`np.array`
+            :class:`~np.array`
 
         :param fluxes:
             An array of fluxes for each star as shape (num_stars, num_pixels).
             The num_stars should match the rows in `labels`.
 
         :type fluxes:
-            :class:`np.ndarray`
+            :class:`~np.ndarray`
 
         :param flux_uncertainties:
             An array of 1-sigma flux uncertainties for each star as shape
@@ -50,14 +54,14 @@ class CannonModel(model.BaseModel):
             should match the `fluxes` array. 
 
         :type flux_uncertainties:
-            :class:`np.ndarray`
+            :class:`~np.ndarray`
         """
 
-        super(CannonModel, self).__init__(labels, fluxes, flux_uncertainties,
-            wavelengths, verify)
+        super(self.__class__, self).__init__(labels, fluxes, flux_uncertainties,
+            wavelengths=wavelengths, verify=verify)
 
 
-    def train(self, label_vector_description, N=None, limits=None, pivot=False,
+    def train(self, label_vector_description, N=None, limits=None, pivot=True,
         **kwargs):
[…]

### 2015-09-07 fireworks/atomic.py
Bugs and improvements

@@ -14,7 +14,7 @@ from oracle import photospheres, synthesis
 logger = logging.getLogger("fireworks")
 
 
-def approximate_atomic_transitions(stellar_parameters, transitions, X_H=False,
+def approximate_atomic_transitions(stellar_parameters, transitions, X_H=True,
     ew_points=None, photosphere_kwargs=None, synthesis_kwargs=None, **kwargs):
     """
     Return functions to approximate the equivalent width of atomic transitions
@@ -214,7 +214,7 @@ def _atomic_line_abundances(stellar_parameters, transitions, X_H=False,
                 "range ({2}, {3})".format(parameter, i, lower, upper))
 
     # Generate values for parameters, where necessary.
-    if ew_points is None: ew_points = np.linspace(10, 300, 30)
+    if ew_points is None: ew_points = np.linspace(10, 300, 15)
     synthesis_kwargs = synthesis_kwargs or {}
     photosphere_kwargs = photosphere_kwargs or {}
 

### 2015-09-07 fireworks/cannon.py

@@ -24,7 +24,7 @@ class CannonModel(model.BaseModel):
         = ("_labels", "_wavelengths", "_fluxes", "_flux_uncertainties")
 
 
-    def __init__(self, labels, wavelengths, fluxes, flux_uncertainties,
+    def __init__(self, labels, fluxes, flux_uncertainties, wavelengths=None,
         verify=True):
         """
         Initialise a Cannon model.
@@ -57,7 +57,7 @@ class CannonModel(model.BaseModel):
             :class:`~np.ndarray`
         """
 
-        super(self.__class__, self).__init__(labels, fluxes, flux_uncertainties,
+        super(CannonModel, self).__init__(labels, fluxes, flux_uncertainties,
             wavelengths=wavelengths, verify=verify)
 
 
@@ -197,7 +197,7 @@ class CannonModel(model.BaseModel):
             if len(term) == 1 and term[0][1] == 1:
                 names.append(term[0][0])
                 indices.append(i)
-        names, indices = tuple(names), np.array(indices)
+        indices = np.array(indices)
 
         if full_output:
             return (indices, names)
@@ -325,6 +325,18 @@ class CannonModel(model.BaseModel):
                 for i, (f, u) in enumerate(zip(flux, flux_uncertainties)):
                     labels[i, :] = _solve_labels_single(f, u, *args, **kwargs)
 
+        return labels
+
+
+    @property
+    @model.requires_training_wheels
+    def lv_labels(self):
+        """
+        Return a list of the labels used to describe the model label vector. 
+        """
+
+        _, labels = self._get_linear_indices(self._label_vector_description,
+            full_output=True)
         return labels
 
 

### 2015-09-07 fireworks/fireworks.py

@@ -17,8 +17,9 @@ logger = logging.getLogger("fireworks")
 
 class FireworksModel(cannon.CannonModel):
 
-    _trained_attributes = ("_coefficients", "_scatter", "_offsets", "_ew_model",
-        "_weak_line_mask", "_label_vector_description", "_atomic_lines")
+    _trained_attributes = ("_coefficients", "_scatter", "_offsets",
+        "_label_vector_description",
+        "_atomic_lines", "_stellar_parameter_labels")
     _data_attributes = ("_labels", "_wavelengths", "_fluxes",
         "_flux_uncertainties")
 
@@ -55,28 +56,29 @@ class FireworksModel(cannon.CannonModel):
             :class:`np.ndarray`
         """
 
-        super(self.__class__, self).__init__(labels, wavelengths, fluxes,
-            flux_uncertainties, verify)
+        super(FireworksModel, self).__init__(labels, fluxes, flux_uncertainties,
+            wavelengths=wavelengths, verify=verify)
 
 
     @property
     @model.requires_training_wheels
-    def labels(self):
+    def lv_labels(self):
         """
         Return a list of the labels involved in this model. This includes any
         labels in the description of the label vector, as well as any individual
         abundance labels.
         """
 
-        _, labels = self._get_linear_indices(self._label_vector_description,
-            full_output=True)
+        labels = super(self.__class__, self).lv_labels
+        # We check for _atomic_lines instead of _trained because we might train
+        # without any atomic lines.
         if self._atomic_lines is not None:
             labels += self._atomic_lines.keys()
         return labels
 
 
-    def train(self, label_vector_description, atomic_lines=None, X_H=False,
-        N=None, limits=None, pivot=False, **kwargs):
+    def train(self, label_vector_description, N=None, limits=None, pivot=False,
+        atomic_lines=None, X_H=True, stellar_parameter_labels=None, **kwargs):
         """
         Train a Cannon model based on the label vector description provided.
 
@@ -85,24 +87,6 @@ class FireworksModel(cannon.CannonModel):
 
         :type label_vector_description:
             str or list of str
-
-        :params atomic_lines: [optional]
-            Atomic absorption lines that should be modelled as part of the first
-            entry in the label vector. If given, this should be a dictionary
-            containing the label names that correspond to $\log_\epsilon(X)$
-            abundances in `CannonModel._labels` as keys, and the values should
-            be an `~astropy.table.Table` containing the atomic transitions for
-            that element.
-
-        :type atomic_lines:
[…]

### 2015-09-07 fireworks/model.py

@@ -160,7 +160,7 @@ class BaseModel(object):
 
 
     @property
-    def labels(self):
+    def lv_labels(self):
         """ This should be overwritten by the parent model. """
         raise WTFError("is this your first day?")
 
@@ -320,7 +320,7 @@ class BaseModel(object):
         # Contents is: trained attributes, data hash, data trained on
         trained_contents = dict(zip(cls._trained_attributes, contents))
         N = len(trained_contents)
-        expected_data_hash = contents[N]
+        expected_data_hash = contents[N-1]
 
         if N + 1 >= len(contents):
             raise TypeError("saved model in {} does not include data".format(
@@ -328,14 +328,14 @@ class BaseModel(object):
 
         # There was data as well.
         if verify and expected_data_hash is not None:
-            actual_data_hash = _short_hash(contents[N + 1:])
+            actual_data_hash = _short_hash(contents[N:])
             if actual_data_hash != expected_data_hash:
                 raise ValueError("expected data hash ({0}) is different ({1})"\
                     .format(expected_data_hash, actual_data_hash))
 
         # Create the model by initialising it with the data attributes.
         model = cls(**dict(zip([_[1:] for _ in cls._data_attributes],
-            contents[N + 1:])))
+            contents[N:])))
 
         # Set the training attributes.
         for k, v in trained_contents.items():

### 2015-09-07 fireworks/fireworks.py
Include zero offsets for atomic lines

@@ -216,6 +216,9 @@ class FireworksModel(cannon.CannonModel):
             ax.plot(self._wavelengths, self._fluxes[0] / weak_line_fluxes[0], c='r', zorder=1000)
             #raise a
 
+            # Update the offsets to be zero for atomic lines.
+            offsets.update({ k: 0 for k in atomic_lines })
+
         else:
             atomic_line_model = None
 
@@ -235,6 +238,7 @@ class FireworksModel(cannon.CannonModel):
 
             if not np.any(np.isfinite(scatter[i] * coefficients[i, :])):
                 logger.warn("No finite coefficients at pixel {}!".format(i))
+
 
         # Save all of these to the model.
         self._trained = True

### 2015-09-07 fireworks/plot.py
Update plotting

@@ -6,8 +6,42 @@
 __author__ = "Andy Casey <arc@ast.cam.ac.uk>"
 
 import numpy as np
-import matplotlib
-import matplotlib.pyplot as plt
+import matplotlib as mpl
+from matplotlib import pyplot as plt
+from matplotlib.ticker import MaxNLocator
+
+cmaps = {
+    "GnBu": [(0.90354479621438422, 0.96276816830915568, 0.88175317890503824),
+             (0.84367551873711977, 0.93903883485233086, 0.82059209066278793),
+             (0.77674741815118231, 0.91252595677095305, 0.76221454704509062),
+             (0.67044984663234042, 0.87118801229140341, 0.71497118192560527),
+             (0.546020769254834,   0.82405229806900027, 0.74740485934650192),
+             (0.41868512595401092, 0.76462900287964763, 0.78985007019603959),
+             (0.29457901909070855, 0.68936564711963433, 0.82066898416070377),
+             (0.19123414667213665, 0.57420994464088881, 0.75866206744137932),
+             (0.092195312678813937, 0.47040370585871677, 0.70579009827445538),
+             (0.031372550874948502, 0.36416763593168822, 0.62755865349489104) ],
+    "RdBu_r": [(0.11864668090699937, 0.37923876006229257, 0.64567475868206403),
+             (0.23660131996753175, 0.54186853357389864, 0.74702039185692293),
+             (0.48143024129026069, 0.7148789097281063, 0.83944637635174923),
+             (0.73241062725291528, 0.8537485669640934, 0.91626298076966228),
+             (0.90142253567190733, 0.93679354471318865, 0.95624759968589335),
+             (0.97923875556272622, 0.91910804019254799, 0.88373703115126667),
+             (0.97970011655022116, 0.78408305785235233, 0.68489044554093303),
+             (0.92226067360709696, 0.56747406545807333, 0.44867361117811766),
+             (0.81153403660830326, 0.32110727271612954, 0.27581700390460445),
+             (0.66920416904430768, 0.084890428419206659, 0.16401384522517523)],
+    "GnBu_d": [(0.21084711176897186, 0.28135332939671537, 0.30823529626808915),
+ (0.22169422353794371, 0.36270665879343067, 0.41647059253617835),
[…]

### 2015-09-07 fireworks/cannon.py
Updates

@@ -362,10 +362,9 @@ class CannonModel(model.BaseModel):
         except AttributeError:
             None
 
-        label_indices, label_names = self._get_linear_indices(
-            self._label_vector_description, full_output=True)
-        expected_labels = np.zeros((self._fluxes.shape[0], len(label_names)))
-        inferred_labels = np.zeros((self._fluxes.shape[0], len(label_names)))
+        names = self.lv_labels
+        expected_labels = np.zeros((self._fluxes.shape[0], len(names)))
+        inferred_labels = np.zeros((self._fluxes.shape[0], len(names)))
 
         # [TODO] Parallelise this?
         # We would need some advice from the user about how many threads to use.
@@ -373,13 +372,13 @@ class CannonModel(model.BaseModel):
         for i, (flux, uncertainty) \
         in enumerate(zip(self._fluxes, self._flux_uncertainties)):
             inferred = self.solve_labels(flux, uncertainty)
-            for j, label_name in enumerate(label_names):
+            for j, label_name in enumerate(names):
                 expected_labels[i, j] = self._labels[label_name][i]
                 inferred_labels[i, j] = inferred[label_name]
 
         # Cache for future, unless the training state changes.
         self._residuals_hash = self._trained_hash
-        self._residuals_cache = (label_names, expected_labels, inferred_labels)
+        self._residuals_cache = (names, expected_labels, inferred_labels)
         return self._residuals_cache
 
 
@@ -477,9 +476,7 @@ class CannonModel(model.BaseModel):
         if label_vector_description is None:
             label_vector_description = self._label_vector_description
 
-        label_indices, label_names = self._get_linear_indices(
-            label_vector_description, full_output=True)
-
+        label_names = self.lv_labels
         N_realisations, N_labels = self._fluxes.shape[0], len(label_names)
         inferred_test_labels = np.nan * np.ones((N_realisations, N_labels))
         expected_test_labels = np.ones((N_realisations, N_labels))
@@ -492,7 +489,6 @@ class CannonModel(model.BaseModel):
             
             mask = np.ones(N_realisations, dtype=bool)
             mask[i] = False
-
 
             # Create a model to use so we don't overwrite self.
             model = self.__class__(self._labels[mask], self._wavelengths,

### 2015-09-07 fireworks/fireworks.py

@@ -402,18 +402,17 @@ class FireworksModel(cannon.CannonModel):
 
         # Create the function.
         def f(coefficients, *labels):
-            # Build the weak lines spectrum.
 
+            # Build the weak lines spectrum.
+            N = len(self._atomic_lines)
+            weak_line_fluxes = np.ones(flux.size)
             stellar_parameters = [labels[self.lv_labels.index(_)] \
                 for _ in self._stellar_parameter_labels]
 
-            weak_line_fluxes = np.ones(flux.size)
-            N = len(self._atomic_lines)
             for i, (label, abundance) \
             in enumerate(zip(self._atomic_lines.keys(), labels[-N:])):
 
                 transitions, ew_coefficients = self._atomic_lines[label]
-
                 for j, mu in enumerate(transitions["wavelength"]):
                     # The 1e-3 factor is to turn the EW from milliAngstroms
                     # into Angstroms.
@@ -446,7 +445,7 @@ class FireworksModel(cannon.CannonModel):
 
         # Remove the temporary finite mask.
         del self._finite_mask
-        
+
         # We might have solved for any number of parameters, so we return a dict
         p_opt = { k: p_opt[i] + self._offsets[k] for i, k in enumerate(names) }
         logger.debug("Final solution: {}".format(p_opt))

### 2015-09-07 fireworks/model.py

@@ -320,7 +320,7 @@ class BaseModel(object):
         # Contents is: trained attributes, data hash, data trained on
         trained_contents = dict(zip(cls._trained_attributes, contents))
         N = len(trained_contents)
-        expected_data_hash = contents[N-1]
+        expected_data_hash = contents[N]
 
         if N + 1 >= len(contents):
             raise TypeError("saved model in {} does not include data".format(
@@ -335,7 +335,7 @@ class BaseModel(object):
 
         # Create the model by initialising it with the data attributes.
         model = cls(**dict(zip([_[1:] for _ in cls._data_attributes],
-            contents[N:])))
+            contents[N+1:])))
 
         # Set the training attributes.
         for k, v in trained_contents.items():

### 2015-09-07 sandbox.py

@@ -189,7 +189,7 @@ atomic_lines = {
 }
 label_vector_description = "TEFF^4 TEFF^3 TEFF^2 TEFF LOGG LOGG^2 TEFF*LOGG TEFF^2*LOGG TEFF*LOGG^2 PARAM_M_H PARAM_M_H*TEFF PARAM_M_H*TEFF^2 PARAM_ALPHA_M PARAM_M_H*PARAM_ALPHA_M" 
 
-
+"""
 temp_filename = "fireworks-Mg+Al+Si+V+Ni.pkl"
 
 import os
@@ -202,6 +202,18 @@ else:
     model = fireworks.FireworksModel.from_filename(temp_filename)
 
 label_names, expected_labels, inferred_labels = model.label_residuals
+"""
+
+"""
+model = fireworks.FireworksModel(stars, wavelengths, fluxes, flux_uncertainties)
+atomic_lines = {
+    "AL_H": AlI_transitions
+}
+result = model.train(label_vector_description, atomic_lines=atomic_lines)
+model.save("t2.pkl", with_data=True, overwrite=True)
+"""
+
+model = fireworks.FireworksModel.from_filename("t2.pkl", verify=False)
 
 
 # 

### 2015-09-07 fireworks/cannon.py
Bugs and parallelise fireworks training

@@ -481,8 +481,7 @@ class CannonModel(model.BaseModel):
         inferred_test_labels = np.nan * np.ones((N_realisations, N_labels))
         expected_test_labels = np.ones((N_realisations, N_labels))
 
-        # Go through each combination.
-        # [TODO] Thread everything.
+        debug = kwargs.get("debug", False)
         for i in range(N_realisations):
             logger.info("Doing cross-validation realisation {0}/{1} on a test "\
                 "set containing 1 star".format(i + 1, N_realisations))
@@ -503,6 +502,7 @@ class CannonModel(model.BaseModel):
             except:
                 logger.exception("Exception in solving star with index {0} in "\
                     "cross-validation".format(i))
+                if debug: raise
 
             else:
                 # Save inferred test labels.

### 2015-09-07 fireworks/fireworks.py

@@ -6,6 +6,7 @@
 __author__ = "Andy Casey <arc@ast.cam.ac.uk>"
 
 import logging
+import multiprocessing as mp
 import numpy as np
 from collections import OrderedDict
 
@@ -132,6 +133,9 @@ class FireworksModel(cannon.CannonModel):
         """
 
 
+        if stellar_parameter_labels is None:
+                stellar_parameter_labels = ["TEFF", "LOGG", "PARAM_M_H"]
+        
         # Since building the atomic line models takes longer than building the
         # label vector array, we build the vector array first so that any errors
         # will appear first.
@@ -172,8 +176,6 @@ class FireworksModel(cannon.CannonModel):
             # We should calculate the expected EWs (and therefore fluxes) for
             # each atomic line for each star in the training set, because this
             # will form the first element of our label vector array.
-            if stellar_parameter_labels is None:
-                stellar_parameter_labels = ["TEFF", "LOGG", "PARAM_M_H"]
             all_stellar_parameters = np.vstack(
                 [self._labels[label] for label in stellar_parameter_labels]).T
 
@@ -204,41 +206,64 @@ class FireworksModel(cannon.CannonModel):
                         weak_line_fluxes[i] *= 1. \
                             - amplitude * np.exp(-(self._wavelengths - mu)**2 \
                                 / (2. * p_sigma**2))
-                    
-
-            import matplotlib.pyplot as plt
-            plt.close("all")
-
-            fig, ax = plt.subplots()
-            ax.plot(self._wavelengths, weak_line_fluxes[0], c='b', zorder=100)
-            ax.plot(self._wavelengths, self._fluxes[0], c='k')
-
-            ax.plot(self._wavelengths, self._fluxes[0] / weak_line_fluxes[0], c='r', zorder=1000)
-            #raise a
 
             # Update the offsets to be zero for atomic lines.
             offsets.update({ k: 0 for k in atomic_lines })
 
+        elif kwargs.get("_atomic_line_model", None):
+            atomic_line_model = kwargs.pop("_atomic_line_model")
+            offsets.update({ k: 0 for k in atomic_line_model })
+
         else:
             atomic_line_model = None
 
-
         assert N is None, "whoops?"
+        N_threads = int(max([1, kwargs.pop("threads", 1)]))
         pb_size = 100 if kwargs.pop("__progressbar", True) else 0
[…]

### 2015-09-07 fireworks/fireworks.py
MAJOR CHANGES

@@ -18,11 +18,11 @@ logger = logging.getLogger("fireworks")
 
 class FireworksModel(cannon.CannonModel):
 
-    _trained_attributes = ("_coefficients", "_scatter", "_offsets",
-        "_label_vector_description",
-        "_atomic_lines", "_stellar_parameter_labels")
-    _data_attributes = ("_labels", "_wavelengths", "_fluxes",
-        "_flux_uncertainties")
+    _trained_attributes \
+        = ("_coefficients", "_scatter", "_offsets", "_label_vector_description",
+            "_stellar_parameter_labels", "_atomic_lines")
+    _data_attributes \
+        = ("_labels", "_wavelengths", "_fluxes", "_flux_uncertainties")
 
     def __init__(self, labels, wavelengths, fluxes, flux_uncertainties,
         verify=True):

### 2015-09-07 article/ms.tex
Jotting down notes

@@
 \newcommand\cv{{\boldsymbol\theta}_\lambda}
+\newcommand\wv{\mathbf{w_n}}
 \newcommand\given{|}
@@
 \begin{abstract}
-More than 15 million spectra are expected to be acquired by stellar surveys in
+More than 15 million stellar spectra are expected to be acquired by surveys in
 the coming decade. The detailed chemical abundance analysis of these data poses
-a significant computational challenge. \tc provides a purely data-driven
+a significant challenge. \tc provides a purely data-driven
 approach to stellar parameter (label) determination. Cross-validation tests
@@
 rely on model atmospheres and spectral synthesis, but \tc is $\sim$300,000 times
-faster, making it perfectly apt for surveys. Simply extending the label vector
-dimensionality of \tc to include individual abundances can create enormous
-model difficulties, severely limiting the interpretability of the results. Here
+faster, making it perfectly applicable for surveys.
+Including individual abundances by extending \tc dimensionality can create enormous model difficulties, severely limiting the interpretability of the results.
+Here
 we describe an alternative approach to including individual element abundances
@@
 impact on model flexibility, allowing for simultaneous training on detailed
-chemical abundances, as well as arbitrary labels where the impact on the spectra
-is unknown or poorly understood (e.g., mass, age). Crucially, we provide a link
-between data-driven models and parallel works on model atmospheres and spectral
-line formation, allowing for data to inform models. We demonstrate that
+chemical abundances (which are formed through well-understood interactions), as well as arbitrary labels where the impact on the spectra
+is unknown or poorly understood (e.g., masses, ages). Crucially, we provide a direct link
+between data-driven models and theory of model atmospheres and spectral line formation. We demonstrate that
 detailed chemical abundances can be measured with a precision of 0.0X-0.0Y~dex
-in Z APOGEE spectra on a modern CPU in just one minute.
+in Z APOGEE spectra on a modern CPU in just one minute. We present abundance
+measurements of 15 elements in Z APOGEE spectra from DR12.
 \end{abstract}
@@
 
+% stellar parameter and chemical abundance analysis take time
+% equation of state, radiative transfer & spectral synthesis of multiple lines
+% 1D model atmospheres, 3D models.
+% introducing tc
 % it's awesome: fast, accurate and precise (given the models published thus far)
 % impressively it does all this whilst ignoring all advances made in stellar physics over the past 100 years.
+% introducing chemical abundances into TC is not trivial
+% high dimensionality
 
@@
 \tc{} provides a purely data-driven approach to stellar label determination. The
-spectral model describes the flux at every pixel $f_{n\lambda}$ to be a function
+spectral model describes the flux $f_{n\lambda}$ for the $n$-th star a pixel with rest-wavelength $\lambda$ to be a function
 of a coefficient vector $\cv$, and a label vector $\lv$:
@@
 
-This form is flexible: the label vector $\lv$ may be linear in just a few labels
-or a complex combination of all $K$ available labels. The underlying principle
-is that the flux for any pixel from any star is predictable with a very simple
-model (generally a polynomial), given the labels. Indeed, \citet{Ness2015}
-employ a quadratic-in-labels (see their Equation 8) spectral model using stellar
-parameters (effective temperature $T_{\rm eff}$, surface gravity $\log{g}$, and 
+This form is flexible: the label vector $\lv$ may be linear in just a few common
+labels (e.g., effective temperature $T_{\rm eff}$, surface gravity $\log{g}$, and
 metallicity [M/H]\footnote{\citet{Ness2015} use [Fe/H] in $\lv$, but for the 
 sake of distinguishing between overall metallicity and individual abundances,
-we opt for [M/H] to describe overall metallicity throughout this \article{}})
-and demonstrate by cross-validation that these labels can be 
+we opt for [M/H] to describe overall metallicity throughout this \article{}} 
+or a complex combination of all $K$ available labels, some of which have an
+ unknown or non-trivial dependence with spectral fluxes \citep[e.g., mass, 
+ ages, see][]{ness_2015b}. The underlying principle
+is that the flux for any pixel from any star is predictable with a simple (usually
+ a polynomial) model, given the labels. Indeed, \citet{Ness2015}
+employ a quadratic-in-labels spectral model using stellar
+parameters ($T_{\rm eff}$, $\log{g}$, [M/H])
+and demonstrate through cross-validation that these labels can be 
 measured with a precision of 95~K in $T_{\rm eff}$, 0.24 in $\log{g}$, 0.08 in 
-[M/H], with negligible biases. It is important to note that these uncertainties
-are comparable to those reported from using \textit{curve-of-growth} on similar 
-data sets (e.g., \citet{X,Y}).
+[M/H], with negligible biases.
 
@@
 photospheres. Calculations of this nature are extremely expensive. Indeed, even
-with simplications of these models (e.g., interpolated 1D photospheres),
+with simplifications of these models (e.g., interpolated 1D photospheres),
 calculating integrated flux requires knowledge of species opacities, detailed 
 atomic and molecular information for all contributing elements, solving for
-the equation of state, and performing radiative transfer calculations. Given the
+the equation of state, and iteratively performing radiative transfer calculations. Given the
 simplicity of \tc{} and a quadratic-in-labels spectral model, it is remarkable
 that these complex calculations can be simplified so much, whilst maintaining
-the typical fidelity reported for high-resolution spectroscopic studies.
+the typical fidelity reported for high-resolution spectroscopic studies based on
+\textit{curve-of-growth} analyses.
+
+% The Cannon is approximating spectral synthesis. Done for R ~ 20k spectra
+% Since all Gaussian and flux value is linearly related to EW, it is reasonable
+% to expect the EW of an individual line can also be approximated with a
+% model like TC.
 
@@
 
-% dealing with integrated EW.
-% APOGEE is R ~ 20,000
+Thus, we have a reasonable form of how EW, log(A) change with respect to
+stellar parameters. At this point we have simply approximated the curve of growth
+for a given line. The coefficients need to be solved for. These can be solved for
+from existing data (fitting profiles) with measured abundances or by synthesising
+spectra and relating the EW to log(X).
+
+% Figure of relating EW to [X/H] with some simple model compared to analytic value.
+
+Consider a scenario where the LSF is known and we want to model what the
+spectrum looks like for three lines. For a given log(X) and stellar parameters,
+we can solve for the EW. If the LSF is known, we can directly turn this into a
+line profile under the assumption of gaussianity.
+
+Multiple (separated) lines can be treated
+
+Include h function.
+
+If lines are blended, one could consider modelling many more atomic transitions.
+However as the number grows, this situation becomes unnecessarily complex:
+often blended lines exist and we do not care about their abundance, only their
+inclusion (nuisance or dirty lines). Other features may be a complex combination
+of thousands of molecular features.
+
+The compromise is to use TC and `clean' lines that we care about in concert.
+
+The spectral model becomes a multiplicative function of $g$ and $h$ with some
+unknown noise.
+
+\begin{equation}
+    f_{n\lambda} = g(\lv\given\cv) \cdot h(\lv,\wv\given\psi_n) +  {\rm noise}.
+\end{equation}
+
+The vector $\wv$ can be calculated directly from spectral synthesis, or using data.
+In practice, given some rest-frame normalised spectra and some knowledge of the
+line spread function, $h(\lv,\wv\given{}X)$ can be calculated before training \tc.
 
@@
 
-\section{Analysis}
-% Detailed chemical abundances with APOGEE.
+\section{APOGEE}
+
+% All about APOGEE Survey and existing data.
+
+% Apply a model using stellar parameters and abundances from X stars in clusters
+% Typical S/N ratio and quality constraints employed
+% Any abundance flags ignored?
+
+% Train a model based on stellar parameters, include XYZ lines (Tabular data)
+
+\section{Discussion}
+
+% Comparison in EW model to log(X)
+% Std deviation in this model.
+
+% Training directly from data, or training from physical models?
+% Inferences that can be made from models (when data are lacking)
+
+% Requirements about knowing the LSF (e.g., a one-parameter LSF is just as good as TC)
+% Blended lines
+
+% Limitations: what about missing abundance data?
 
@@
 
+%
+
 \section*{Acknowledgments}
+This research can be reproduced in full using the commands \texttt{pip install fireworks},
+\texttt{fireworks-reproduce}. This article was produced using commit hash \texttt{hash}.
+
 This research was fostered by discussions at the \textit{New Milky Way}
@@
 
+\section{Appendix A: Reproducing this research}
+This article has been compiled from the `git` repository hosted online at GitHub (commit hash `x` on `master` branch).
+The figures and tables included in this article can be reproduced in full using the following commands in a modern terminal:
+
+git clone git@github.com:andycasey/fireworks.git fireworks
+cd fireworks
+python setup.py install
+python reproduce.py
+cd article
+make
+
+If the required tools are not available on the readers system (e.g., `git`), this research can also be produced using the public Docker image .
+
 \end{document}

### 2015-09-08 article/ms.tex
Draft; awful; don't read

@@
 \newcommand\given{|}
+
+\newcommand\teff{$T_{\rm eff}$}
+\newcommand\logg{$\log{g}$}
+\newcommand\mh{${\mathrm [M/H]}$}
 
@@
 \begin{abstract}
-More than 15 million stellar spectra are expected to be acquired by surveys in
+More than 15 million stellar spectra are expected to be acquired by ground-based surveys in
 the coming decade. The detailed chemical abundance analysis of these data poses
@@
 \section{Introduction}
+\label{sec:introduction}
+
+Ground-based stellar spectroscopic surveys have enormous scientific potential.
+Key absorption features in stellar spectra permit inferences on the internal structure of stars and their detailed chemical composition.
+Given a large enough sample of stars, these inferences can directly guide or constrain models over a large range of physical scales: the formation of planets, stars and clusters, supernovae yields and nucleosynthetic pathways, as well as the large scale evolution of the Milky Way.
+For these reasons an increasing number of stellar spectroscopic surveys are being undertaken, with each survey generation increasing in sample size \citep{gaia_eso; galah; lamost; apogee; 4most}.
+
+% Data have outgrown the methods.
+In some ways, the spectroscopic data have outgrown the methods used to analyse them.
+High-resolution (R > 20,000) stellar spectroscopy is colloquially considered to be a `dark art': analyses are carefully performed by an expert spectroscopist, who themselves were trained by an expert in exactly where the stellar continuum should be placed, and which atomic lines are considered 'clean' or 'dirty'.
+This work is equally impressive as it is subjective: the best spectroscopists host an encyclopaedic knowledge of atomic absorption lines with decades of experience, yet in blind-tests their decisions of where to place the continuum quickly becomes the most dominant source of systematic error.
+% Lot of unextracted information in these spectra (e.g., mass, ages)
+
+% Chemical abundance analysis takes time.
+Classical spectroscopic analyses are time-intensive both in human and computing resources.
+The Gaia-ESO Survey has been particularly instrumental in minimising the human resources required for analysis: all analysis codes of FGK stars employed in Gaia-ESO are now completely automatised.
+The computing challenge remains: some analysis codes take up to one hour (single core) to determine stellar parameters (\teff, \logg, \mh) for one star.
+Given the data scale (e.g., $\sim$400 stars observed simultaneously with HERMES), this poses a modest computational problem.
+
+% These analysis techniques are not optimal, either.
+These analyses also have strict limitations in validity, and a range of options are available.
+The community standard is to perform chemical abundance analyses using 1D model photospheres under the assumption of local thermal equilibrium (LTE).
+There are a number of photospheric models available.
+Detailed information about the species (atomic and molecular) opacities is required, and partition functions are usually approximated from polynomials.
+Atomic data (e.g., oscillator strengths) must be well-known, and the treatment of Stark and XX broadening must be carefully included.
+Results are further sensitive to the manner in which the equation of state was solved.
+Simply put, there are many things to consider for a careful analysis.
+
+% TC is in stark contrast to these methods
+% It's fast, accurate and precise.
+
+% It ignores physics.
+\tc{} falls into the regime of supervised machine learning.
+The results obtained by \tc{} thus far are extremely impressive, particularly when it is considered that the model ignores everything learned about atomic physics and stellar spectra since the names' inconception 100 years ago.
+There is no knowledge of stars, no input on interactions between fields and particles, nothing considering spectral line formation, and no input of atomic data, opacities, partition functions or depth-dependent photospheric quantities.
+Only a polynomial model of stellar parameter labels is required in order to reasonably approximate expensive spectral syntheses.
+However because \tc is purely data-driven, there is a missing link to physics.
+
+% Including chemical abundances into TC is non-trivial.
 
@@
 sake of distinguishing between overall metallicity and individual abundances,
-we opt for [M/H] to describe overall metallicity throughout this \article{}} 
-or a complex combination of all $K$ available labels, some of which have an
- unknown or non-trivial dependence with spectral fluxes \citep[e.g., mass, 
+we opt for [M/H] to describe overall metallicity throughout this \article{}})
+or a complex combination of all $K$ available labels, some of which may have an
+ unknown or non-trivial relationship with spectral fluxes \citep[e.g., mass, 
  ages, see][]{ness_2015b}. The underlying principle
@@
 with simplifications of these models (e.g., interpolated 1D photospheres),
-calculating integrated flux requires knowledge of species opacities, detailed 
+calculating synthetic spectra requires knowledge of species opacities, detailed 
 atomic and molecular information for all contributing elements, solving for
-the equation of state, and iteratively performing radiative transfer calculations. Given the
-simplicity of \tc{} and a quadratic-in-labels spectral model, it is remarkable
-that these complex calculations can be simplified so much, whilst maintaining
-the typical fidelity reported for high-resolution spectroscopic studies based on
+the equation of state, and iteratively performing expensive radiative transfer
+calculations. Given the
+simplicity of a quadratic-in-labels spectral model with \tc{}, it is remarkable
+that these complex calculations can be simplified so greatly, whilst maintaining
+the typical measurement fidelity reported for high-resolution spectroscopic studies based on
 \textit{curve-of-growth} analyses.
+
+In a way, \tc{} is approximating spectral synthesis. Thus far the applications
+have been limited to stellar spectra with spectral resolving powers of 20,000
+or less. In this regime, the profiles of weak (linearly-responding) spectral lines
+are well-represented by Gaussian profiles. In
 

### 2015-09-08 sandbox.py
Test

@@ -188,7 +188,6 @@ atomic_lines = {
 
 }
 label_vector_description = "TEFF^4 TEFF^3 TEFF^2 TEFF LOGG LOGG^2 TEFF*LOGG TEFF^2*LOGG TEFF*LOGG^2 PARAM_M_H PARAM_M_H*TEFF PARAM_M_H*TEFF^2 PARAM_ALPHA_M PARAM_M_H*PARAM_ALPHA_M" 
-
 """
 temp_filename = "fireworks-Mg+Al+Si+V+Ni.pkl"
 
@@ -201,6 +200,7 @@ if not os.path.exists(temp_filename):
 else:
     model = fireworks.FireworksModel.from_filename(temp_filename)
 
+print("Done")
 label_names, expected_labels, inferred_labels = model.label_residuals
 """
 

### 2015-09-08 article/ms.tex
Test

@@
 \usepackage{amsmath,amsbsy}
+\usepackage{gitinfo2}
 
@@
 \section{Appendix A: Reproducing this research}
-This article has been compiled from the `git` repository hosted online at GitHub (commit hash `x` on `master` branch).
+This article has been compiled from the `git` repository hosted online at GitHub (commit hash \gitHash on `master` branch).
 The figures and tables included in this article can be reproduced in full using the following commands in a modern terminal:

### 2015-09-08 article/ms.tex
Test hooks
Include git abbreviated hash, etc whoohoo

@@
 
+% Comment
+
 % Convenient colloquiualisms.
-\newcommand\article{\textit Article}
+\newcommand\article{\textit{Article}}
 \newcommand\tc{\textit{The Cannon} }
@@
 \section*{Acknowledgments}
-This research can be reproduced in full using the commands \texttt{pip install fireworks},
-\texttt{fireworks-reproduce}. This article was produced using commit hash \texttt{hash}.
-
 This research was fostered by discussions at the \textit{New Milky Way}
@@
 \section{Appendix A: Reproducing this research}
-This article has been compiled from the `git` repository hosted online at GitHub (commit hash \gitHash on `master` branch).
-The figures and tables included in this article can be reproduced in full using the following commands in a modern terminal:
+This \article{} has been compiled from the \texttt{git} repository (commit hash \texttt{\gitAbbrevHash} on \texttt{\gitBranch} branch) hosted online at GitHub\footnote{https://www.github.com/andycasey/fireworks}.
+The complete data analysis, figures and tables included in this \article{} can be reproduced in full using the following commands in a modern Unix-based terminal:
+\vspace{0.5em}
 
-git clone git@github.com:andycasey/fireworks.git fireworks
-cd fireworks
-python setup.py install
-python reproduce.py
-cd article
-make
-
-If the required tools are not available on the readers system (e.g., `git`), this research can also be produced using the public Docker image .
+\noindent\texttt{[reproducibility commands]} 
+\vspace{0.5em} \\
+If the required tools are not available on the readers system (e.g., \texttt{git}), this research can also be produced using the public Docker image \texttt{[docker image reference]}.
 

### 2015-09-08 article/ms.tex
Minor updates

@@
 \section{Appendix A: Reproducing this research}
-This \article{} has been compiled from the \texttt{git} repository (commit hash \texttt{\gitAbbrevHash} on \texttt{\gitBranch} branch) hosted online at GitHub\footnote{https://www.github.com/andycasey/fireworks}.
+This \article{} has been compiled from the \texttt{git} repository on \gitAuthorDate{} (commit hash \texttt{\gitAbbrevHash} on \texttt{\gitBranch} branch) hosted online at GitHub\footnote{https://www.github.com/andycasey/fireworks}.
 The complete data analysis, figures and tables included in this \article{} can be reproduced in full using the following commands in a modern Unix-based terminal:

### 2015-09-08 fireworks/atomic.py
Predictions go through .predict now (and bug fixes)

@@ -345,6 +345,8 @@ def _approximate_radiative_transfer(transition, stellar_parameters, ews,
 
     return (p_opt, p_cov)
 
+
+
 """
 
 def _tile_transition_values(values, N_stars, N_ews):

### 2015-09-08 fireworks/cannon.py

@@ -15,6 +15,7 @@ from . import (model, plot, utils)
 
 logger = logging.getLogger("fireworks")
 
+from time import time
 
 class CannonModel(model.BaseModel):
 
@@ -369,6 +370,7 @@ class CannonModel(model.BaseModel):
         # [TODO] Parallelise this?
         # We would need some advice from the user about how many threads to use.
 
+        t = time()
         for i, (flux, uncertainty) \
         in enumerate(zip(self._fluxes, self._flux_uncertainties)):
             inferred = self.solve_labels(flux, uncertainty)
@@ -376,6 +378,7 @@ class CannonModel(model.BaseModel):
                 expected_labels[i, j] = self._labels[label_name][i]
                 inferred_labels[i, j] = inferred[label_name]
 
+        print("DONE ", i, time() - t)
         # Cache for future, unless the training state changes.
         self._residuals_hash = self._trained_hash
         self._residuals_cache = (names, expected_labels, inferred_labels)

### 2015-09-08 fireworks/fireworks.py

@@ -15,12 +15,15 @@ from . import (atomic, cannon, model, plot, utils)
 
 logger = logging.getLogger("fireworks")
 
+# TODO
+from time import time
+
 
 class FireworksModel(cannon.CannonModel):
 
     _trained_attributes \
         = ("_coefficients", "_scatter", "_offsets", "_label_vector_description",
-            "_stellar_parameter_labels", "_atomic_lines")
+            "_atomic_lines", "_stellar_parameter_labels")
     _data_attributes \
         = ("_labels", "_wavelengths", "_fluxes", "_flux_uncertainties")
 
@@ -78,7 +81,7 @@ class FireworksModel(cannon.CannonModel):
         return labels
 
 
-    def train(self, label_vector_description, N=None, limits=None, pivot=False,
+    def train(self, label_vector_description, N=None, limits=None, pivot=True,
         atomic_lines=None, X_H=True, stellar_parameter_labels=None, **kwargs):
         """
         Train a Cannon model based on the label vector description provided.
@@ -311,23 +314,26 @@ class FireworksModel(cannon.CannonModel):
                 "given: {2}".format(len(names), len(labels),
                     ", ".join(names)))
 
+        t_i = time()
         # Generate the Cannon-ical flux.
         label_vector_indices = self._parse_label_vector_description(
             self._label_vector_description, return_indices=True,
             __columns=names)
 
-        offsets = np.array([self._offsets[names] for names in namess])
+        offsets = np.array([self._offsets[name] for name in names])
         fluxes = np.dot(self._coefficients, cannon._build_label_vector_rows(
             label_vector_indices, labels - offsets).T).flatten()
 
+        t_a = time()
+        tc = 0
         # Include treatment of any atomic lines.
         if self._atomic_lines is not None:
 
             N = len(self._atomic_lines)
-            stellar_parameters = [labels[labels.index(_)] \
+            stellar_parameters = [labels[names.index(_)] \
                 for _ in self._stellar_parameter_labels]
             
-            weak_line_fluxes = np.ones(cannonical_flux.size)
+            #weak_line_fluxes = np.ones(fluxes.size)
             for i, (label, abundance) \
             in enumerate(zip(self._atomic_lines.keys(), labels[-N:])):
 
@@ -335,20 +341,25 @@ class FireworksModel(cannon.CannonModel):
 
                 for j, mu in enumerate(transitions["wavelength"]):
[…]

### 2015-09-08 article/ms.tex
Added more description about a basic cannon model

@@
 
-\usepackage{amsmath,amsbsy}
+\usepackage{amsmath,amsfonts,amssymb,amsbsy}
 \usepackage{gitinfo2}
-
-% Comment
 
@@
 \newcommand\article{\textit{Article}}
-\newcommand\tc{\textit{The Cannon} }
+\newcommand\tc{\textit{The Cannon}}
 
@@
 \newcommand\wv{\mathbf{w_n}}
+\newcommand\ssq{s_\lambda^2}
 \newcommand\given{|}
+\DeclareMathOperator*{\argmax}{\arg\!\max}
 
@@
 
-\title[Cannon Chemistry]{Detailed Chemical Abundances with \tc} 
+\title[Cannon Chemistry]{Detailed Chemical Abundances with \tc{}} 
 % Some arrangement of the following authors:
@@
 the coming decade. The detailed chemical abundance analysis of these data poses
-a significant challenge. \tc provides a purely data-driven
+a significant challenge. \tc{} provides a purely data-driven
 approach to stellar parameter (label) determination. Cross-validation tests
 demonstrate that the precision achievable is comparable to existing methods that
-rely on model atmospheres and spectral synthesis, but \tc is $\sim$300,000 times
+rely on model atmospheres and spectral synthesis, but \tc{} is $\sim$300,000 times
 faster, making it perfectly applicable for surveys.
-Including individual abundances by extending \tc dimensionality can create enormous model difficulties, severely limiting the interpretability of the results.
+Including individual abundances by extending \tc{} dimensionality can create enormous model difficulties, severely limiting the interpretability of the results.
 Here
 we describe an alternative approach to including individual element abundances
-in \tc. Using modest simplifications of the physics of spectral line formation,
-we demonstrate precisely \textit{why} an approach like \tc works so effectively,
+in \tc{}. Using modest simplifications of the physics of spectral line formation,
+we demonstrate precisely \textit{why} an approach like \tc{} works so effectively,
 and provide a physically-motivated model that describes flux values, given the
@@
 In some ways, the spectroscopic data have outgrown the methods used to analyse them.
-High-resolution (R > 20,000) stellar spectroscopy is colloquially considered to be a `dark art': analyses are carefully performed by an expert spectroscopist, who themselves were trained by an expert in exactly where the stellar continuum should be placed, and which atomic lines are considered 'clean' or 'dirty'.
+High-resolution ($\mathcal{R} > 20,000$) stellar spectroscopy is colloquially considered to be a `dark art': analyses are carefully performed by an expert spectroscopist, who themselves were trained by an expert in exactly where the stellar continuum should be placed, and which atomic lines are considered 'clean' or 'dirty'.
 This work is equally impressive as it is subjective: the best spectroscopists host an encyclopaedic knowledge of atomic absorption lines with decades of experience, yet in blind-tests their decisions of where to place the continuum quickly becomes the most dominant source of systematic error.
@@
 Only a polynomial model of stellar parameter labels is required in order to reasonably approximate expensive spectral syntheses.
-However because \tc is purely data-driven, there is a missing link to physics.
+However because \tc{} is purely data-driven, there is a missing link to physics.
 
@@
 \tc{} provides a purely data-driven approach to stellar label determination. The
-spectral model describes the flux $f_{n\lambda}$ for the $n$-th star a pixel with rest-wavelength $\lambda$ to be a function
+spectral model describes the flux $f_{n\lambda}$ (for star $n$) at a pixel with rest-wavelength $\lambda$ to be a function
 of a coefficient vector $\cv$, and a label vector $\lv$:
@@
     f_{n\lambda} = g(\lv\given\cv) + {\rm noise}.
+    \label{eq:cannon}
 \end{equation}
@@
 This form is flexible: the label vector $\lv$ may be linear in just a few common
-labels (e.g., effective temperature $T_{\rm eff}$, surface gravity $\log{g}$, and
-metallicity [M/H]\footnote{\citet{Ness2015} use [Fe/H] in $\lv$, but for the 
+labels (e.g., \teff, \logg, \mh\footnote{\citet{Ness15a} use [Fe/H] in $\lv$, but for the 
 sake of distinguishing between overall metallicity and individual abundances,
-we opt for [M/H] to describe overall metallicity throughout this \article{}})
+throughout this \article{} we opt for \mh to describe overall metallicity})
 or a complex combination of all $K$ available labels, some of which may have an
- unknown or non-trivial relationship with spectral fluxes \citep[e.g., mass, 
- ages, see][]{ness_2015b}. The underlying principle
+ unknown or non-trivial relationship with spectral fluxes \citep[e.g., ages, 
+ masses, see][]{Nissen15, Ness15b}. The underlying principle
 is that the flux for any pixel from any star is predictable with a simple (usually
- a polynomial) model, given the labels. Indeed, \citet{Ness2015}
+ a polynomial) model, given the labels. Indeed, \citet{Ness15a}
 employ a quadratic-in-labels spectral model using stellar
@@
 
-Utilising a simple polynomial model to predict spectral fluxes is a distinctive
-method from parallel efforts in radiative transfer of realistic (3D) stellar 
-photospheres. Calculations of this nature are extremely expensive. Indeed, even
-with simplifications of these models (e.g., interpolated 1D photospheres),
-calculating synthetic spectra requires knowledge of species opacities, detailed 
-atomic and molecular information for all contributing elements, solving for
-the equation of state, and iteratively performing expensive radiative transfer
-calculations. Given the
-simplicity of a quadratic-in-labels spectral model with \tc{}, it is remarkable
-that these complex calculations can be simplified so greatly, whilst maintaining
-the typical measurement fidelity reported for high-resolution spectroscopic studies based on
-\textit{curve-of-growth} analyses.
+The noise in Equation \ref{eq:cannon} can be taken as a quadrature combination of the pixel uncertainty variance $\sigma_{n\lambda}^2$ and the intrinsic variance of the spectral model at each wavelength $s_\lambda^2$.
+The flux $f_{n\lambda}$ can be written as a linear function of the label vector, leading to a single-pixel log-likelihood function
 
-In a way, \tc{} is approximating spectral synthesis. Thus far the applications
+\begin{equation}
+\ln{p}\left(f_{n\lambda}\given\cv^\intercal,\lv, s_{\lambda}^2\right) = -\frac{1}{2}\frac{\left[f_{n\lambda} - \cv^{\intercal}\cdot\lv\right]^2}{s_\lambda^2 + \sigma_{n\lambda}^2} -\frac{1}{2}\ln\left(s_{\lambda}^2 + \sigma_{n\lambda}^2\right).
+\label{eq:pixel-log-likelihood}
+\end{equation}
+
+Given some sample of stars with known labels (the training set), the label vector $\lv$ and fluxes $f_{n\lambda}$ are known, allowing us to solve for the parameters $[\cv,s_\lambda^2]$ for each pixel with wavelength $\lambda$.
+Here we will optimise the negative log-likelihood (Equation \ref{eq:pixel-log-likelihood}).
+This is the training step of \tc{}, where the parameters $[\cv,s_\lambda^2]$ are optimised for a single pixel at $\lambda$ by considering all stars $N$:
+
+\begin{equation}
+\cv,s_\lambda \leftarrow^{\arg\,\max} {\cv, s_\lambda} \sum_{n=1}^N \ln{p}\left(f_{n\lambda}\given\cv^\intercal,\lv,s_\lambda^2\right)
+\end{equation}
+
+Once the coefficients $[\cv,\ssq]$ are known, \tc{} provides a \textit{generative model} for spectral fluxes.
+Through a cheap linear product of $\cv^\intercal$ and $\lv$, \tc{} is approximating spectral synthesis, an otherwise extremely expensive endeavour.
+We refer the reader to \citet{Ness15a} for a more detailed description of \tc{}.
+
+Before extending \tc{} to include individual chemical abundances, it is
+a useful exercise to understand \textit{why} a simple spectral model can 
+approximate synthesis calculations with comparable fidelity.
+Thus far the applications of \tc{} have been restricted to stellar spectra with resolving powers of $\mathcal{R} = 22,500$ \citep[APOGEE;][]{Ness15a} or less \citep[e.g., LAMOST;][]{Ho15}.
+In these regimes, the shapes of unsaturated spectral lines are well-represented by Gaussian profiles.
+
+% The reason
+%extends past model flexibility: one could imagine a similar coefficient vector
+%$\cv$ with a different functional form of the label vector $\lv$ (e.g., an
+%extreme case might be $\lv \equiv [1, {\rm [M/H]}^{T_{\rm eff}},
+%\log{g}^{T_{\rm eff}}]$) which would probably be incapable of describing
+%spectral fluxes with any fidelity.
+
+% Consider a single line.
+
+% This is why we should also be able to model the EW->log(X) with a Cannon-like model.
+
+% However this is complex because if we extended this to infinity, we would need an infinity of blended lines, etc.
+
+%
+
+%Utilising a simple polynomial model to predict spectral fluxes is a distinctive
+%method from parallel efforts in radiative transfer of realistic (3D) stellar 
+%photospheres. Calculations of this nature are extremely expensive. Indeed, even
+%with simplifications of these models (e.g., interpolated 1D photospheres),
+%calculating synthetic spectra requires knowledge of species opacities, detailed 
+%atomic and molecular information for all contributing elements, solving for
+%the equation of state, and iteratively performing expensive radiative transfer
+%calculations. Given the
+%simplicity of a quadratic-in-labels spectral model with \tc{}, it is remarkable
+%that these complex calculations can be simplified so greatly, whilst maintaining
+%the typical measurement fidelity reported for high-resolution spectroscopic studies based on
+%\textit{curve-of-growth} analyses.
+
+% Solving for coefficients and scatter term.
+
+Thus far the applications
 have been limited to stellar spectra with spectral resolving powers of 20,000
 or less. In this regime, the profiles of weak (linearly-responding) spectral lines
-are well-represented by Gaussian profiles. In
+are well-represented by Gaussian profiles.
 
@@
 
-Before extending this framework to include individual chemical abundances, it is
-a useful exercise to understand \textit{why} a simple polynomial model can 
-approximate spectral synthesis calculations with comparable fidelity. The reason
-extends past model flexibility: one could imagine a similar coefficient vector
-$\cv$ with a different functional form of the label vector $\lv$ (e.g., an
-extreme case might be $\lv \equiv [1, {\rm [M/H]}^{T_{\rm eff}},
-\log{g}^{T_{\rm eff}}]$) which would probably be incapable of describing
-spectral fluxes with any fidelity.
+\textbf{Description of the model of EW formation, and how this changes as a function of teff, logg, mh}
 
-% Describe the model of EW formation
-
-% How this changes as a function of teff, logg, feh
+We now have a functional form for how $W$ and [X/H] (or $\log_{\epsilon}(X)$) are related, given some stellar parameters, which themselves are part of the set of $K$ labels.
 
@@
 
-\bibitem[Ness et al.(2015)]{Ness2015} Ness, M., Hogg, D.~W., 
-Rix, H.-W., Ho, A.~Y.~Q., \& Zasowski, G.\ 2015, \apj, 808, 16 
+\bibitem[Ho et al.(2015)]{Ho15} Ho, A., et al., in preparation
+\bibitem[Ness et al.(2015a)]{Ness15a} Ness, M., Hogg, D.~W., Rix, H.-W., Ho, A.~Y.~Q., \& Zasowski, G.\ 2015, \apj, 808, 16
+
+\bibitem[Ness et al.(2015b)]{Ness15b} Ness, M., et al., in preparation
+
+\bibitem[Nissen(2015)]{Nissen15} Nissen, P.~E.\ 2015, \aap, 579, A52
+
 \end{thebibliography}

### 2015-09-08 article/ms.tex
Introduce EW

@@
 
-The noise in Equation \ref{eq:cannon} can be taken as a quadrature combination of the pixel uncertainty variance $\sigma_{n\lambda}^2$ and the intrinsic variance of the spectral model at each wavelength $s_\lambda^2$.
+The noise (Eq.
+\ref{eq:cannon}) can be taken as a quadrature combination of the pixel uncertainty variance $\sigma_{n\lambda}^2$ and the intrinsic variance of the spectral model at each wavelength $s_\lambda^2$.
 The flux $f_{n\lambda}$ can be written as a linear function of the label vector, leading to a single-pixel log-likelihood function
@@
 a useful exercise to understand \textit{why} a simple spectral model can 
-approximate synthesis calculations with comparable fidelity.
-Thus far the applications of \tc{} have been restricted to stellar spectra with resolving powers of $\mathcal{R} = 22,500$ \citep[APOGEE;][]{Ness15a} or less \citep[e.g., LAMOST;][]{Ho15}.
+approximate synthesis calculations with comparable fidelity, and thus \textit{how} this framework can be extended to include individual abundances.
+At present the applications of \tc{} have been restricted to stellar spectra with resolving powers of $\mathcal{R} = 22,500$ \citep[APOGEE;][]{Ness15a} or less \citep[e.g., LAMOST;][]{Ho15}.
 In these regimes, the shapes of unsaturated spectral lines are well-represented by Gaussian profiles.
+This fact leads to several critical insights, which we describe below.
+
+In stellar spectroscopy we are frequently presented with the idea of the equivalent width (EW; $\mathcal{W}$) of an absorption line.
+Consider a single (unblended) atomic line at rest wavelength $\lambda_r$, where the flux values at and surrounding $\lambda_r$ ($f_{n\lambda_r-{\delta}},f_{n\lambda_{r+\delta}}$) are well-approximated by a \textit{Cannon} spectral model.
+Because the absorption line is well-represented by a Gaussian profile at this spectral resolution, the flux at $\lambda_r$ is related to the EW of the line through
+
+\begin{equation}
+\mathcal{W} = \sigma\cdot{}f_{n\lambda_r}\sqrt{2\pi}
+\end{equation}
+
+where $\sigma$ is the standard deviation of the profile, which is well-approximated by width of the line spread function (LSF) kernel.
+That is to say, for resolutions $\mathcal{R} \lesssim 20,000$, $\sigma$ is dominated by the spectrograph, not the absorption line or the labels (stellar parameters).
+
+% where the flux values $f_{n\lambda}$ are well-approximated by the given spectral model.
+
+% Things are Gaussian, therefore EW related.
+% How does EW of a line change w.r.t. stellar parameters
+% If we had a model for a single line, how could this be included.
 
@@
 Astropy, a community-developed core Python package for Astronomy \citep{astropy}.
+The authors are thankful for the following free online services which help make
+science reproducibility easier: GitHub, Travis Continuous Integration, and Zenodo.
 
@@
 \section{Appendix A: Reproducing this research}
-This \article{} has been compiled from the \texttt{git} repository on \gitAuthorDate{} (commit hash \texttt{\gitAbbrevHash} on \texttt{\gitBranch} branch) hosted online at GitHub\footnote{https://www.github.com/andycasey/fireworks}.
-The complete data analysis, figures and tables included in this \article{} can be reproduced in full using the following commands in a modern Unix-based terminal:
+This \article{} has been compiled from the \texttt{git} repository (abbreviated commit hash \texttt{\gitAbbrevHash} on \texttt{\gitBranch} branch) hosted online by GitHub at \texttt{github.com/andycasey/fireworks}.
+The data required to reproduce this analysis is persistently hosted through Zenodo \citep{DATA}.
+The complete analysis, figures and tables included in this \article{} can be reproduced in full using the following commands in a modern Unix-based terminal:
 \vspace{0.5em}

### 2015-09-08 article/ms.tex
Adding text

@@
 
+This \article{} is organised as follows.
+In Section \ref{sec:model} we briefly describe the existing foundations for \tc{} and demonstrate from a theoretical perspective as to why the model flexibility is sufficient to accurately represent spectral fluxes.
+We also outline how to include individual atomic lines within an existing \textit{Cannon} model.
+In Section \ref{sec:apogee} we apply a detailed chemical abundance \textit{Cannon} to the APOGEE Data Release (DR) 12 spectra.
+We discuss the suitability and limitations of the method in Section \ref{sec:discussion}.
+In Section \ref{sec:conclusion} we conclude with a summary of our findings.
+
 \section{Model}
+\label{sec:model}
 
@@
 \begin{equation}
-\mathcal{W} = \sigma\cdot{}f_{n\lambda_r}\sqrt{2\pi}
+\mathcal{W} = \sigma{}\sqrt{2\pi}(1 - f_{n\lambda_r})
+\label{eq:ew-flux}
 \end{equation}
@@
 That is to say, for resolutions $\mathcal{R} \lesssim 20,000$, $\sigma$ is dominated by the spectrograph, not the absorption line or the labels (stellar parameters).
+If we presume that the LSF is well-known, or reasonably approximated by a fixed spectral resolution $\mathcal{R}$, then it is clear to see that \tc{} is implicitly modelling the EW of an atomic line as a function of stellar labels, and modelling spectral flux through a per-pixel scaling of $\sigma$.
+Indeed, the success of \tc{} suggests that the EW of an atomic line can also be well-approximated by a simple model relating the labels (i.e., stellar parameters and abundance of a single species).
+If such a model can be constructed and appropriately included within the existing \textit{Cannon} framework, it would permit the determination of detailed chemical abundances, whilst maintaining the flexibility in training on labels with poorly-known relationships on spectral fluxes (e.g., ages).
 
-% where the flux values $f_{n\lambda}$ are well-approximated by the given spectral model.
+We note that Equation \ref{eq:ew-flux} does not imply that \tc{} is limited in applicability to low-resolution ($\mathcal{R} \lesssim 20,000$) spectra.
+Quite the opposite.
+\tc{} is sufficiently flexible that in extremely high-resolution spectra, where absorption lines are more closely represented by Lorentzian or Hertzting functions, the flux values $f_{n\lambda}$ in the wings of a strong line are presumably still well-modelled by \tc{}.
+Indeed, the line $\mathcal{W}$ would still be predictable with high-fidelity, implying that the coefficients $\cv$ for pixels in the wings just scale in a different way.
+Equation \ref{eq:ew-flux} allows us to directly translate $\mathcal{W}$ to $f_{n\lambda}$, thereby accounting for the flux contribution from a single atomic line.
+If the spectral resolving power were sufficiently high such that the absorption profiles appeared Lorentzian in nature, some prior information on the shape $\gamma$ would be required to translate $\mathcal{W}$ to $f_{n\lambda}$.
+Thus, while Equation \ref{eq:ew-flux} allows the translation from $\mathcal{W}$ to $f_{n\lambda}$ for $\mathcal{R} \lesssim 20,000$ spectra, given some knowledge of the LSF, no limiting interpretation about \tc{} should be drawn.
+Importantly, all future stellar spectroscopic surveys  currently expect $\mathcal{R} \lesssim 20,000$, in the perfect Gaussian-profile regime for this extension of \tc{}.
 
-% Things are Gaussian, therefore EW related.
+% Show an example?
+The example could have a figure showing some spectral fluxes for a clean line, which is modelled by the cannon (with an example model spectrum) and then show the same model being used to train on EW.
+%An example of this is shown in Figure \ref{fig:cannon-ew-relation}, where a quadratic-in-labels \textit{Cannon} model is used to model spectral fluxes for X stars across the HR diagram.
+The data have been trained
+
+Knowing that a simple model can relate the a single line abundance [X/H] and integrated strength of an atomic line $\mathcal{W}$, given stellar parameter labels, it is intriguing to ask whether model atmosphere relations and the theory of spectral line formation can guide the level of model complexity required.
+
 % How does EW of a line change w.r.t. stellar parameters
-% If we had a model for a single line, how could this be included.
+\textbf{Description of how EW for a single line will change w.r.t stellar parameters}
 
-% The reason
-%extends past model flexibility: one could imagine a similar coefficient vector
-%$\cv$ with a different functional form of the label vector $\lv$ (e.g., an
-%extreme case might be $\lv \equiv [1, {\rm [M/H]}^{T_{\rm eff}},
-%\log{g}^{T_{\rm eff}}]$) which would probably be incapable of describing
-%spectral fluxes with any fidelity.
+We now have a functional form of a model that relates $\mathcal{W}$ to $\log_\epsilon\left({\textrm X}\right)$ (or a scaled-solar [X/H] value), given the stellar parameters, which themselves are part of the set of $K$ labels.
+It is interesting to note that Equation \ref{eq:ew-logx-model} is not dissimilar to the quadratic-in-labels model used by \citet{Ness15a} for APOGEE data.
+With only the prior knowledge that \textit{a simple model of the labels can predict spectral fluxes}, we have derived the appropriate functional form for a single absorption line.
 
-% Consider a single line.
-
-% This is why we should also be able to model the EW->log(X) with a Cannon-like model.
-
+% How do we include this atomic line model.
 % However this is complex because if we extended this to infinity, we would need an infinity of blended lines, etc.
-
-%
 
@@
 % Solving for coefficients and scatter term.
-
-Thus far the applications
-have been limited to stellar spectra with spectral resolving powers of 20,000
-or less. In this regime, the profiles of weak (linearly-responding) spectral lines
-are well-represented by Gaussian profiles.
-
-% The Cannon is approximating spectral synthesis. Done for R ~ 20k spectra
-% Since all Gaussian and flux value is linearly related to EW, it is reasonable
-% to expect the EW of an individual line can also be approximated with a
-% model like TC.
-
-\textbf{Description of the model of EW formation, and how this changes as a function of teff, logg, mh}
-
-We now have a functional form for how $W$ and [X/H] (or $\log_{\epsilon}(X)$) are related, given some stellar parameters, which themselves are part of the set of $K$ labels.
 

### 2015-09-13 article/ms.tex
Bad words

@@
 
-% Convenient colloquiualisms.
 \newcommand\article{\textit{Article}}
@@
 
-% Convenient maths.
 \newcommand\K{$\mathbf{K}$}
@@
 With only the prior knowledge that \textit{a simple model of the labels can predict spectral fluxes}, we have derived the appropriate functional form for a single absorption line.
+The model $\mathcal{W}\left({\rm [X/H]}, T, logg\right)$ simply approximates the curve-of-growth for a single line, and the coefficients still need to be calculated.
+The practicalities of determining these coefficients is outlined in Section \ref{sec:somewhere}.
+For the purposes of introducing the model, we can assume that they are easily calculable.
 
-% How do we include this atomic line model.
-% However this is complex because if we extended this to infinity, we would need an infinity of blended lines, etc.
+Consider multiple atomic lines, where the coefficients to translate $\mathcal{W}$ and [X/H] are known.
+Given some knowledge of the line spread function, the EWs $\mathcal{W}$ to flux values $f_{n\lambda}$ can be translated such that:
+
+\begin{equation}
+h = X
+\label{eq:h-model}
+\end{equation}
+
+Here we implicitly assume the mutual opacity contributions of nearby weak lines will be small. 
+At this point it is worthwhile to ask whether Equation \ref{eq:h-model} can be scaled sufficiently to $M$ atomic lines in order to produce an entire spectral model for $f_{n\lambda}$, without the contribution of $g\left(\lv\given\cv\right)$.
+There would be a number of immediate issues with that approach: a huge number of additional coefficients would be required, hidden lines could not be encapsulated accurately by the model, strong lines would be extremely poorly modelled by the fit, no longer linear (slow), but most importantly the pixels would lose flexibility in training on arbitrary labels.
+
+If lines are blended, one could consider modelling many more atomic transitions.
+However as the number grows, this situation becomes unnecessarily complex:
+often blended lines exist and we do not care about their abundance, only their
+inclusion (nuisance or dirty lines). Other features may be a complex combination
+of thousands of molecular features.
+
+Here we introduce a method of including individual atomic transitions within an existing \textit{Cannon} model.
+
+The spectral model becomes a multiplicative function of $g$ and $h$ with some
+unknown noise.
+
+\begin{equation}
+    f_{n\lambda} = g(\lv\given\cv) \cdot h(\lv,\wv\given\psi_n) +  {\rm noise}.
+\end{equation}
+
+This inclusion combines the existing flexibility in the model, allowing for training on arbitrary labels, and uses individual absorption lines for which some information is known (species, atomic data, etc) in order to trace the chemical abundances of individual elements.
+
+The vector $\wv$ can be calculated directly from spectral synthesis, or using data.
+In practice, given some rest-frame normalised spectra and some knowledge of the
+line spread function, $h(\lv,\wv\given{}X)$ can be calculated before training \tc.
 
@@
 
-% Solving for coefficients and scatter term.
-
-Thus, we have a reasonable form of how EW, log(A) change with respect to
-stellar parameters. At this point we have simply approximated the curve of growth
-for a given line. The coefficients need to be solved for. These can be solved for
-from existing data (fitting profiles) with measured abundances or by synthesising
-spectra and relating the EW to log(X).
-
 % Figure of relating EW to [X/H] with some simple model compared to analytic value.
 
-Consider a scenario where the LSF is known and we want to model what the
-spectrum looks like for three lines. For a given log(X) and stellar parameters,
-we can solve for the EW. If the LSF is known, we can directly turn this into a
-line profile under the assumption of gaussianity.
-
-Multiple (separated) lines can be treated
-
-Include h function.
-
-If lines are blended, one could consider modelling many more atomic transitions.
-However as the number grows, this situation becomes unnecessarily complex:
-often blended lines exist and we do not care about their abundance, only their
-inclusion (nuisance or dirty lines). Other features may be a complex combination
-of thousands of molecular features.
-
 The compromise is to use TC and `clean' lines that we care about in concert.
-
-The spectral model becomes a multiplicative function of $g$ and $h$ with some
-unknown noise.
-
-\begin{equation}
-    f_{n\lambda} = g(\lv\given\cv) \cdot h(\lv,\wv\given\psi_n) +  {\rm noise}.
-\end{equation}
-
-The vector $\wv$ can be calculated directly from spectral synthesis, or using data.
-In practice, given some rest-frame normalised spectra and some knowledge of the
-line spread function, $h(\lv,\wv\given{}X)$ can be calculated before training \tc.
 

### 2015-10-06 article/ms.tex
Mostly word-wrapping, with some minor text changes

@@
 \begin{abstract}
-More than 15 million stellar spectra are expected to be acquired by ground-based surveys in
+More than 15 million stellar spectra will be acquired by ground-based surveys in
 the coming decade. The detailed chemical abundance analysis of these data poses
@@
 Ground-based stellar spectroscopic surveys have enormous scientific potential.
-Key absorption features in stellar spectra permit inferences on the internal structure of stars and their detailed chemical composition.
-Given a large enough sample of stars, these inferences can directly guide or constrain models over a large range of physical scales: the formation of planets, stars and clusters, supernovae yields and nucleosynthetic pathways, as well as the large scale evolution of the Milky Way.
-For these reasons an increasing number of stellar spectroscopic surveys are being undertaken, with each survey generation increasing in sample size \citep{gaia_eso; galah; lamost; apogee; 4most}.
+  Key absorption features in stellar spectra permit inferences on the internal 
+structure of stars and their detailed chemical composition.  Given a sufficiently
+large sample of stars, these inferences can directly guide or constrain models 
+over a large range of physical scales: the formation of planets, stars and 
+clusters, supernovae yields and nucleosynthetic pathways, as well as the large 
+scale evolution of the Milky Way.  For these reasons an increasing number of 
+stellar spectroscopic surveys are being undertaken, with each survey generation 
+increasing in sample size \citep{gaia_eso; galah; lamost; apogee; 4most}.
 
 % Data have outgrown the methods.
-In some ways, the spectroscopic data have outgrown the methods used to analyse them.
-High-resolution ($\mathcal{R} > 20,000$) stellar spectroscopy is colloquially considered to be a `dark art': analyses are carefully performed by an expert spectroscopist, who themselves were trained by an expert in exactly where the stellar continuum should be placed, and which atomic lines are considered 'clean' or 'dirty'.
-This work is equally impressive as it is subjective: the best spectroscopists host an encyclopaedic knowledge of atomic absorption lines with decades of experience, yet in blind-tests their decisions of where to place the continuum quickly becomes the most dominant source of systematic error.
+In some ways, the spectroscopic data have outgrown the methods used to analyse
+them. High-resolution ($\mathcal{R} > 20,000$) stellar spectroscopy has 
+colloquially been considered as a `dark art': analyses are carefully performed 
+by an expert spectroscopist, who themselves were trained by an expert in exactly
+where the stellar continuum should be placed, and which atomic lines are 
+considered `clean' or `dirty'. This work is equally impressive as it is 
+subjective: the best spectroscopists host an encyclopaedic knowledge of atomic
+absorption lines with decades of experience, yet in blind-tests their decisions 
+of where to place the continuum quickly becomes the most dominant source of 
+systematic error.
 % Lot of unextracted information in these spectra (e.g., mass, ages)
@@
 % Chemical abundance analysis takes time.
-Classical spectroscopic analyses are time-intensive both in human and computing resources.
-The Gaia-ESO Survey has been particularly instrumental in minimising the human resources required for analysis: all analysis codes of FGK stars employed in Gaia-ESO are now completely automatised.
-The computing challenge remains: some analysis codes take up to one hour (single core) to determine stellar parameters (\teff, \logg, \mh) for one star.
-Given the data scale (e.g., $\sim$400 stars observed simultaneously with HERMES), this poses a modest computational problem.
+Classical spectroscopic analyses are time-intensive both in human and computing
+resources. The Gaia-ESO Survey has been particularly instrumental in minimising
+the human resources required for analysis: all analysis codes of FGK stars 
+employed in Gaia-ESO are now completely automatised.  However the computing
+challenge remains: some analysis codes take up to one hour (single core) to 
+determine stellar parameters (\teff, \logg, \mh) for a single star. Given the 
+data scale (e.g., $\sim$400 stars observed simultaneously with HERMES), this 
+poses a modest computational problem.
 
 % These analysis techniques are not optimal, either.
-These analyses also have strict limitations in validity, and a range of options are available.
-The community standard is to perform chemical abundance analyses using 1D model photospheres under the assumption of local thermal equilibrium (LTE).
-There are a number of photospheric models available.
-Detailed information about the species (atomic and molecular) opacities is required, and partition functions are usually approximated from polynomials.
-Atomic data (e.g., oscillator strengths) must be well-known, and the treatment of Stark and XX broadening must be carefully included.
-Results are further sensitive to the manner in which the equation of state was solved.
-Simply put, there are many things to consider for a careful analysis.
+These analyses also have strict limitations in validity, and a range of options
+are available.  Many of these options are not straightforward.  The community 
+standard is to perform chemical abundance analyses using 1D model photospheres 
+under the assumption of local thermal equilibrium (LTE). There are a number of 
+photospheric models available.  Detailed information about the species (atomic 
+and molecular) opacities is required, and partition functions are usually 
+approximated from polynomials. Atomic data (e.g., oscillator strengths) must be
+well-known, and the treatment of Stark and XX broadening must be carefully
+included. Results are further sensitive to the manner in which the equation of 
+state was solved. Simply put, there are many things to consider for a careful 
+analysis.
 
@@
 This form is flexible: the label vector $\lv$ may be linear in just a few common
-labels (e.g., \teff, \logg, \mh\footnote{\citet{Ness15a} use [Fe/H] in $\lv$, but for the 
-sake of distinguishing between overall metallicity and individual abundances,
-throughout this \article{} we opt for \mh to describe overall metallicity})
-or a complex combination of all $K$ available labels, some of which may have an
- unknown or non-trivial relationship with spectral fluxes \citep[e.g., ages, 
- masses, see][]{Nissen15, Ness15b}. The underlying principle
-is that the flux for any pixel from any star is predictable with a simple (usually
- a polynomial) model, given the labels. Indeed, \citet{Ness15a}
-employ a quadratic-in-labels spectral model using stellar
-parameters ($T_{\rm eff}$, $\log{g}$, [M/H])
-and demonstrate through cross-validation that these labels can be 
-measured with a precision of 95~K in $T_{\rm eff}$, 0.24 in $\log{g}$, 0.08 in 
-[M/H], with negligible biases.
+labels (e.g., \teff, \logg, \mh\footnote{\citet{Ness15a} use [Fe/H] in $\lv$, 
+but for the sake of distinguishing between overall metallicity and individual 
+abundances, throughout this \article{} we opt for \mh to describe overall 
+metallicity}) or a complex combination of all $K$ available labels, some of which
+may have an unknown or non-trivial relationship with spectral fluxes \citep[e.g.,
+ages, masses, see][]{Nissen15, Ness15b}. The underlying principle is that the 
+flux for any pixel from any star is predictable with a simple model, given the
+labels. Indeed, \citet{Ness15a} employ a quadratic-in-labels spectral model 
+using stellar parameters ($T_{\rm eff}$, $\log{g}$, [M/H]) and demonstrate 
+through cross-validation that these labels can be measured with a precision of 
+95~K in $T_{\rm eff}$, 0.24 in $\log{g}$, 0.08 in [M/H], with negligible biases.
 
-The noise (Eq.
-\ref{eq:cannon}) can be taken as a quadrature combination of the pixel uncertainty variance $\sigma_{n\lambda}^2$ and the intrinsic variance of the spectral model at each wavelength $s_\lambda^2$.
-The flux $f_{n\lambda}$ can be written as a linear function of the label vector, leading to a single-pixel log-likelihood function
+% TODO: This needs rewording:
+The noise in Equation \ref{eq:cannon} can be taken as a quadrature combination
+of the variance in pixel flux $\sigma_{n\lambda}^2$ and the intrinsic variance 
+of the spectral model at each wavelength $s_\lambda^2$. The flux $f_{n\lambda}$
+can be written as a linear function of the label vector, leading to the
+single-pixel log-likelihood function
 
@@
 
-Given some sample of stars with known labels (the training set), the label vector $\lv$ and fluxes $f_{n\lambda}$ are known, allowing us to solve for the parameters $[\cv,s_\lambda^2]$ for each pixel with wavelength $\lambda$.
-Here we will optimise the negative log-likelihood (Equation \ref{eq:pixel-log-likelihood}).
-This is the training step of \tc{}, where the parameters $[\cv,s_\lambda^2]$ are optimised for a single pixel at $\lambda$ by considering all stars $N$:
+The parameters $[\cv,s_\lambda^2]$ for each pixel with wavelength $\lambda$ need
+to be inferred. This is accomplished given some sample of stars where the labels
+are known with some fidelity (i.e., the training set). Here we will optimise the
+negative log-likelihood (Equation \ref{eq:pixel-log-likelihood}). This is the 
+training step of \tc{}, where the parameters $[\cv,s_\lambda^2]$ are optimised 
+for a single pixel at $\lambda$ by considering all stars $N$:
 
+% TODO: Need to re-place 'arg max'
 \begin{equation}
@@
 
-Once the coefficients $[\cv,\ssq]$ are known, \tc{} provides a \textit{generative model} for spectral fluxes.
-Through a cheap linear product of $\cv^\intercal$ and $\lv$, \tc{} is approximating spectral synthesis, an otherwise extremely expensive endeavour.
-We refer the reader to \citet{Ness15a} for a more detailed description of \tc{}.
+Once the coefficients $\{\cv,\ssq\}$ are known, \tc{} provides an approximate
+\textit{generative model} for spectral fluxes. Through a cheap linear product of
+$\cv^\intercal$ and $\lv$, \tc{} is approximating spectral synthesis, an 
+otherwise extremely expensive endeavour. This provides a number of advantages,
+the most straightforward of which is the ability to efficiently determine $\lv$
+(i.e., the stellar parameters) for a new star outside the training set in a
+fraction of a second, which is about five orders of magnitude faster than
+existing analysis codes. We refer the reader to \citet{Ness15a} for a more
+detailed description of \tc{}.
 
-Before extending \tc{} to include individual chemical abundances, it is
-a useful exercise to understand \textit{why} a simple spectral model can 
-approximate synthesis calculations with comparable fidelity, and thus \textit{how} this framework can be extended to include individual abundances.
-At present the applications of \tc{} have been restricted to stellar spectra with resolving powers of $\mathcal{R} = 22,500$ \citep[APOGEE;][]{Ness15a} or less \citep[e.g., LAMOST;][]{Ho15}.
-In these regimes, the shapes of unsaturated spectral lines are well-represented by Gaussian profiles.
-This fact leads to several critical insights, which we describe below.
+It is not straightforward to add individual chemical abundances as labels and
+expect equally impressive results.  Some chemical abundances only affect few
+pixels of the spectrum. Other elements only present transitions that are
+frequently blended with other elements, implying covariance between elemental
+abundances.  And all transitions will have linear (or higher-order) 
+relationships with stellar parameters.  Addressing these issues by increasing
+the complexity of the label vector introduces additional issues (e.g.,
+over-fitting, interpretability).  Although the existing \textit{Cannon}
+framework can be extended to include individual chemical abundances, they cannot
+be introduced by extensions of the label vector without additional issues.
 
-In stellar spectroscopy we are frequently presented with the idea of the equivalent width (EW; $\mathcal{W}$) of an absorption line.
-Consider a single (unblended) atomic line at rest wavelength $\lambda_r$, where the flux values at and surrounding $\lambda_r$ ($f_{n\lambda_r-{\delta}},f_{n\lambda_{r+\delta}}$) are well-approximated by a \textit{Cannon} spectral model.
-Because the absorption line is well-represented by a Gaussian profile at this spectral resolution, the flux at $\lambda_r$ is related to the EW of the line through
+% TODO: rewrite this so it all flows better.
+Before we extend \tc{} to include individual chemical abundances, it is a useful
+ exercise to understand \textit{why} a simple spectral model can 
+approximate synthesis calculations with comparable fidelity, and thus 
+\textit{how} this framework can be extended to include individual abundances. At
+present the applications of \tc{} have been restricted to stellar spectra with 
+resolving powers of $\mathcal{R} = 22,500$ \citep[APOGEE;][]{Ness15a} or less 
+\citep[e.g., LAMOST;][]{Ho15}. In these regimes, the shapes of unsaturated 
+spectral lines are well-represented by Gaussian profiles. This leads to several 
+crucial insights, which we describe below.
+
+In stellar spectroscopy we are frequently presented with the idea of the 
+equivalent width (EW; $\mathcal{W}$) of a spectral line. Consider a single 
+(unblended) atomic line at rest wavelength $\lambda_r$, where the flux values 
+surrounding $\lambda_r$ ($f_{n\lambda_r-{\delta}},f_{n\lambda_{r+\delta}}$) are 
+well-approximated by a \textit{Cannon} spectral model. Because the absorption 
+line is well-represented by a Gaussian profile at this spectral resolution, the 
+flux at $\lambda_r$ is related to the EW of the line through
 
@@
 
-where $\sigma$ is the standard deviation of the profile, which is well-approximated by width of the line spread function (LSF) kernel.
-That is to say, for resolutions $\mathcal{R} \lesssim 20,000$, $\sigma$ is dominated by the spectrograph, not the absorption line or the labels (stellar parameters).
-If we presume that the LSF is well-known, or reasonably approximated by a fixed spectral resolution $\mathcal{R}$, then it is clear to see that \tc{} is implicitly modelling the EW of an atomic line as a function of stellar labels, and modelling spectral flux through a per-pixel scaling of $\sigma$.
-Indeed, the success of \tc{} suggests that the EW of an atomic line can also be well-approximated by a simple model relating the labels (i.e., stellar parameters and abundance of a single species).
-If such a model can be constructed and appropriately included within the existing \textit{Cannon} framework, it would permit the determination of detailed chemical abundances, whilst maintaining the flexibility in training on labels with poorly-known relationships on spectral fluxes (e.g., ages).
+where $\sigma$ is the standard deviation of the profile. For typical FGK-type
+stars observed with $\mathcal{R} \lesssim 20,000$, $\sigma$ will be dominated by
+the line spread function (LSF). If we presume that the LSF is well-known, or 
+reasonably approximated by a fixed spectral resolution $\mathcal{R}$, then it is
+clear to see that \tc{} is implicitly modelling the EW of an atomic line as a 
+function of stellar labels, and modelling spectral flux through a per-pixel 
+scaling of $\sigma$. Indeed, the success of \tc{} suggests that the EW of an 
+atomic line can also be well-approximated by a simple model relating the labels 
+(i.e., stellar parameters and abundance of a single species). If such a model 
+could be constructed and appropriately included within the existing 
+\textit{Cannon} framework, it would permit the determination of individual 
+chemical abundances, whilst maintaining the flexibility in training on labels 
+with poorly-known relationships on spectral fluxes (e.g., ages).
 
-We note that Equation \ref{eq:ew-flux} does not imply that \tc{} is limited in applicability to low-resolution ($\mathcal{R} \lesssim 20,000$) spectra.
-Quite the opposite.
-\tc{} is sufficiently flexible that in extremely high-resolution spectra, where absorption lines are more closely represented by Lorentzian or Hertzting functions, the flux values $f_{n\lambda}$ in the wings of a strong line are presumably still well-modelled by \tc{}.
-Indeed, the line $\mathcal{W}$ would still be predictable with high-fidelity, implying that the coefficients $\cv$ for pixels in the wings just scale in a different way.
-Equation \ref{eq:ew-flux} allows us to directly translate $\mathcal{W}$ to $f_{n\lambda}$, thereby accounting for the flux contribution from a single atomic line.
-If the spectral resolving power were sufficiently high such that the absorption profiles appeared Lorentzian in nature, some prior information on the shape $\gamma$ would be required to translate $\mathcal{W}$ to $f_{n\lambda}$.
-Thus, while Equation \ref{eq:ew-flux} allows the translation from $\mathcal{W}$ to $f_{n\lambda}$ for $\mathcal{R} \lesssim 20,000$ spectra, given some knowledge of the LSF, no limiting interpretation about \tc{} should be drawn.
-Importantly, all future stellar spectroscopic surveys  currently expect $\mathcal{R} \lesssim 20,000$, in the perfect Gaussian-profile regime for this extension of \tc{}.
+We note that Equation \ref{eq:ew-flux} does not imply that \tc{} is limited in
+applicability to low-resolution ($\mathcal{R} \lesssim 20,000$) spectra. Quite 
+the opposite. \tc{} is sufficiently flexible that in extremely high-resolution 
+spectra, where absorption lines are more closely represented by Lorentzian or
+Hertzting functions, the flux values $f_{n\lambda}$ in the wings of a strong 
+line are still well-modelled by \tc{} (as judged by $\ssq$ and performance in
+cross-validation tests). Indeed, the line $\mathcal{W}$ would still be 
+predictable with high-fidelity, implying that the coefficients $\cv$ for pixels
+in the wings just scale in a different way. Equation \ref{eq:ew-flux} allows us
+to directly translate $\mathcal{W}$ to $f_{n\lambda}$, thereby accounting for 
+the flux contribution from a single atomic line. If the spectral resolving power
+were sufficiently high such that the absorption profiles appeared Lorentzian in
+nature, some prior information on the shape $\gamma$ would be required to 
+translate $\mathcal{W}$ to $f_{n\lambda}$. Thus, while Equation \ref{eq:ew-flux}
+allows the translation from $\mathcal{W}$ to $f_{n\lambda}$ for 
+$\mathcal{R} \lesssim 20,000$ spectra, given some knowledge of the LSF, no 
+limiting interpretation about \tc{} should be drawn. Importantly, all future 
+stellar spectroscopic surveys are currently planning for spectral resolutions
+$\mathcal{R} \lesssim 20,000$, in the perfect Gaussian-profile regime for this
+extension of \tc{}.
 

### 2015-10-27 article/ms.tex
Some braindumping
Cleaned up a lot of the model text

@@
 \newcommand\logg{$\log{g}$}
-\newcommand\mh{${\mathrm [M/H]}$}
+\newcommand\mh{${\rm [M/H]}$}
+\newcommand{\set}[1]{$\{#1\}$}
+\newcommand{\logeps}[1]{\log_{\epsilon}{\rm #1}}
+\newcommand\ew{$\mathcal{W}$}
+\newcommand\psiv{\mathbf{\boldsymbol{\psi}_m}}
 
-\title[Cannon Chemistry]{Detailed Chemical Abundances with \tc{}} 
+\title[Detailed Chemical Abundances with \tc{}]{Detailed Chemical Abundances with \tc{}}
+
 % Some arrangement of the following authors:
@@
 
-\date{Accepted 2015 XX XX. Received 2015 October XX; in original form 2015 XX xx}
+\date{Accepted 2015 XX XX. Received 2015 XX XX; in original form 2015 XX XX}
 
@@
 structure of stars and their detailed chemical composition.  Given a sufficiently
-large sample of stars, these inferences can directly guide or constrain models 
+large sample of stars, these inferences can exclude, guide, or constrain models 
 over a large range of physical scales: the formation of planets, stars and 
 clusters, supernovae yields and nucleosynthetic pathways, as well as the large 
-scale evolution of the Milky Way.  For these reasons an increasing number of 
-stellar spectroscopic surveys are being undertaken, with each survey generation 
-increasing in sample size \citep{gaia_eso; galah; lamost; apogee; 4most}.
+scale evolution of the Milky Way.  For these reasons, an increasing number of 
+stellar spectroscopic surveys are being undertaken, each 
+increasing in sample size \citep{gaia_eso, galah, lamost, apogee, weave, 4most}.
 
-% Data have outgrown the methods.
-In some ways, the spectroscopic data have outgrown the methods used to analyse
-them. High-resolution ($\mathcal{R} > 20,000$) stellar spectroscopy has 
-colloquially been considered as a `dark art': analyses are carefully performed 
+We propose the spectroscopic data have outgrown the methods used to analyse
+them. There exists an inordinate amount of high quality (high-resolution, high S/N)
+spectra, far more than what was available just a decade ago, however the techniques 
+employed to extract information from these data have improved only fractionally. 
+A subjective (human-driven) analysis was manageable given the data volume, and 
+for this reason high-resolution ($\mathcal{R} > 20,000$) stellar spectroscopy has 
+colloquially been considered as a `dark art'. Analyses were carefully performed 
 by an expert spectroscopist, who themselves were trained by an expert in exactly
-where the stellar continuum should be placed, and which atomic lines are 
-considered `clean' or `dirty'. This work is equally impressive as it is 
+where the stellar continuum ought to be placed, which atomic lines are 
+considered `clean' or `dirty' (blended), as well as which analysis decisions are
+acceptable, and which are not. This work is equally impressive as it is 
 subjective: the best spectroscopists host an encyclopaedic knowledge of atomic
@@
 of where to place the continuum quickly becomes the most dominant source of 
-systematic error.
-% Lot of unextracted information in these spectra (e.g., mass, ages)
+systematic error \citep{who, who}.
 
-% Chemical abundance analysis takes time.
-Classical spectroscopic analyses are time-intensive both in human and computing
-resources. The Gaia-ESO Survey has been particularly instrumental in minimising
-the human resources required for analysis: all analysis codes of FGK stars 
-employed in Gaia-ESO are now completely automatised.  However the computing
-challenge remains: some analysis codes take up to one hour (single core) to 
-determine stellar parameters (\teff, \logg, \mh) for a single star. Given the 
-data scale (e.g., $\sim$400 stars observed simultaneously with HERMES), this 
-poses a modest computational problem.
+Classical spectroscopic analyses are human time-intensive, and were unable to
+scale to the current data volume. The Gaia-ESO Survey has been particularly 
+influential in minimising the human resources required for analysis: all analysis 
+codes of FGK stars employed in Gaia-ESO are completely automatised.  However
+a computing challenge has arisen: some analysis codes take up to one hour (single
+core) to determine stellar parameters (\teff, \logg, \mh) for a single star. Given the 
+expected data growth in the next decade, this poses a major computational error.
 
-% These analysis techniques are not optimal, either.
 These analyses also have strict limitations in validity, and a range of options
-are available.  Many of these options are not straightforward.  The community 
-standard is to perform chemical abundance analyses using 1D model photospheres 
-under the assumption of local thermal equilibrium (LTE). There are a number of 
-photospheric models available.  Detailed information about the species (atomic 
-and molecular) opacities is required, and partition functions are usually 
-approximated from polynomials. Atomic data (e.g., oscillator strengths) must be
-well-known, and the treatment of Stark and XX broadening must be carefully
-included. Results are further sensitive to the manner in which the equation of 
-state was solved. Simply put, there are many things to consider for a careful 
-analysis.
+are available.  Many analysis options do not have straightforward answers. The 
+community standard is to perform chemical abundance analyses using a 
+one-dimensional description of a steady-state model photosphere with the
+assumption of local thermal equilibrium (LTE). There are a number of 
+photospheric models available from different groups.  Detailed information about
+the species (atomic and molecular) opacities is required, and partition functions
+are usually approximated from polynomials published decades earlier. Atomic 
+data (e.g., oscillator strengths) must be well-known, and the treatment of 
+natural, thermal and pressure broadening must be appropriately included,
+with careful consideration to the impact on perturber collisions. Results are 
+further sensitive to the manner in which the equation of state was solved. 
+Simply put, there is a lot of physics to consider.
 
-% TC is in stark contrast to these methods
-% It's fast, accurate and precise.
+In contrast, \tc{} \citep{Ness15a} ignores everything we know about physics.
+\tc{} falls into the regime of supervised machine learning, and is $\sim10^5$
+times faster than standard spectroscopic analysis codes. The results published 
+using \tc{} thus far are extremely impressive. There is no knowledge of 
+stars, no input on interactions between electromagnetic fields and particles, nothing considering
+spectral line formation, and no input of atomic data, opacities, partition functions 
+or depth-dependent photospheric quantities. Only a relatively small training set
+and a polynomial model of stellar parameter labels is required in order to 
+reasonably approximate expensive spectral syntheses. However because 
+\tc{} is purely data-driven, there is a missing link to physics. Most crucially,
+because \tc{} works in label space, including individual chemical abundances
+is a non-trivial exercise. Massive increases in the label complexity would be
+required to account for known covariances between parameters and elements,
+which increases the likelihood of overfitting and limits robustness and interpretability.
+Given the high quality spectra currently being collected by ground-based surveys,
+extending functionality of \tc\ to include individual chemical abundances -- whilst
+maintaining interpretability -- has the potential to revolutionise the way we 
+treat stellar spectroscopic data.
 
-% It ignores physics.
-\tc{} falls into the regime of supervised machine learning.
-The results obtained by \tc{} thus far are extremely impressive, particularly when it is considered that the model ignores everything learned about atomic physics and stellar spectra since the names' inconception 100 years ago.
-There is no knowledge of stars, no input on interactions between fields and particles, nothing considering spectral line formation, and no input of atomic data, opacities, partition functions or depth-dependent photospheric quantities.
-Only a polynomial model of stellar parameter labels is required in order to reasonably approximate expensive spectral syntheses.
-However because \tc{} is purely data-driven, there is a missing link to physics.
-
-% Including chemical abundances into TC is non-trivial.
-
-% stellar parameter and chemical abundance analysis take time
-% equation of state, radiative transfer & spectral synthesis of multiple lines
-% 1D model atmospheres, 3D models.
-% introducing tc
-% it's awesome: fast, accurate and precise (given the models published thus far)
-% impressively it does all this whilst ignoring all advances made in stellar physics over the past 100 years.
-% introducing chemical abundances into TC is not trivial
-% high dimensionality
-
-This \article{} is organised as follows.
-In Section \ref{sec:model} we briefly describe the existing foundations for \tc{} and demonstrate from a theoretical perspective as to why the model flexibility is sufficient to accurately represent spectral fluxes.
-We also outline how to include individual atomic lines within an existing \textit{Cannon} model.
-In Section \ref{sec:apogee} we apply a detailed chemical abundance \textit{Cannon} to the APOGEE Data Release (DR) 12 spectra.
+This \article{} is organised as follows. In Section \ref{sec:model} we briefly describe
+the existing foundations for \tc{} and demonstrate from a theoretical perspective why
+the model flexibility is sufficient to accurately represent spectral fluxes. This 
+introduction is crucial to outlining how we later include individual atomic lines within 
+an existing \textit{Cannon} model. In Section \ref{sec:apogee} we apply a detailed
+chemical abundance \textit{Cannon} to the APOGEE Data Release (DR) 12 spectra.
 We discuss the suitability and limitations of the method in Section \ref{sec:discussion}.
@@
 but for the sake of distinguishing between overall metallicity and individual 
-abundances, throughout this \article{} we opt for \mh to describe overall 
-metallicity}) or a complex combination of all $K$ available labels, some of which
+abundances, throughout this \article{} we opt for \mh\ to describe overall 
+metallicity.}) or a complex combination of all $K$ available labels, some of which
 may have an unknown or non-trivial relationship with spectral fluxes \citep[e.g.,
-ages, masses, see][]{Nissen15, Ness15b}. The underlying principle is that the 
+ages and masses, see][]{Nissen15, Ness15b}. The underlying principle is that the 
 flux for any pixel from any star is predictable with a simple model, given the
@@
 through cross-validation that these labels can be measured with a precision of 
-95~K in $T_{\rm eff}$, 0.24 in $\log{g}$, 0.08 in [M/H], with negligible biases.
+95~K in $T_{\rm eff}$, 0.24 in $\log{g}$, and 0.08 in [M/H] with negligible biases.
 
-% TODO: This needs rewording:
 The noise in Equation \ref{eq:cannon} can be taken as a quadrature combination
 of the variance in pixel flux $\sigma_{n\lambda}^2$ and the intrinsic variance 
-of the spectral model at each wavelength $s_\lambda^2$. The flux $f_{n\lambda}$
-can be written as a linear function of the label vector, leading to the
-single-pixel log-likelihood function
+of the spectral model at each wavelength $s_\lambda^2$, which, like $\cv$, is
+not known before the model is trained (see below).
+
+The flux $f_{n\lambda}$ can then be written as a linear function of the label 
+vector, leading to the single-pixel log-likelihood function
 
 \begin{equation}
-\ln{p}\left(f_{n\lambda}\given\cv^\intercal,\lv, s_{\lambda}^2\right) = -\frac{1}{2}\frac{\left[f_{n\lambda} - \cv^{\intercal}\cdot\lv\right]^2}{s_\lambda^2 + \sigma_{n\lambda}^2} -\frac{1}{2}\ln\left(s_{\lambda}^2 + \sigma_{n\lambda}^2\right).
+\ln{p}\left(f_{n\lambda}\given\cv^\intercal,\lv, s_{\lambda}\right) = -\frac{1}{2}\frac{\left[f_{n\lambda} - \cv^{\intercal}\cdot\lv\right]^2}{s_\lambda^2 + \sigma_{n\lambda}^2} -\frac{1}{2}\ln\left(s_{\lambda}^2 + \sigma_{n\lambda}^2\right).
 \label{eq:pixel-log-likelihood}
@@
 
-The parameters $[\cv,s_\lambda^2]$ for each pixel with wavelength $\lambda$ need
-to be inferred. This is accomplished given some sample of stars where the labels
-are known with some fidelity (i.e., the training set). Here we will optimise the
-negative log-likelihood (Equation \ref{eq:pixel-log-likelihood}). This is the 
-training step of \tc{}, where the parameters $[\cv,s_\lambda^2]$ are optimised 
+The parameters \set{\cv,s_\lambda^2} for each pixel with wavelength $\lambda$
+need to be inferred. This is accomplished given some sample of stars where 
+the labels are known with high fidelity (i.e., the training set). Here we will 
+optimise the negative log-likelihood (Equation \ref{eq:pixel-log-likelihood}),
+which we later confirm to be sufficient for our purposes. This is the training 
+step of \tc{}, where the parameters \set{\cv,s_\lambda^2} are optimised 
 for a single pixel at $\lambda$ by considering all stars $N$:
 
-% TODO: Need to re-place 'arg max'
 \begin{equation}
-\cv,s_\lambda \leftarrow^{\arg\,\max} {\cv, s_\lambda} \sum_{n=1}^N \ln{p}\left(f_{n\lambda}\given\cv^\intercal,\lv,s_\lambda^2\right)
+\cv,s_\lambda \leftarrow \substack{\mbox{arg\,max}\\{\{\cv , s_\lambda\}}} \sum_{n=1}^N \ln{p}\left(f_{n\lambda}\given\cv^\intercal,\lv,s_\lambda^2\right)
 \end{equation}
 
-Once the coefficients $\{\cv,\ssq\}$ are known, \tc{} provides an approximate
-\textit{generative model} for spectral fluxes. Through a cheap linear product of
-$\cv^\intercal$ and $\lv$, \tc{} is approximating spectral synthesis, an 
-otherwise extremely expensive endeavour. This provides a number of advantages,
+Once the pixel coefficients and variance \set{\cv, s_\lambda^2} are known,
+\tc{} provides an approximate \textit{generative model} for spectral fluxes. 
+Through a cheap linear product of $\cv^\intercal$ and $\lv$, \tc{} is 
+effectively (and efficiently) approximating spectral synthesis, an otherwise
+extremely expensive endeavour. This provides a number of advantages,
 the most straightforward of which is the ability to efficiently determine $\lv$
-(i.e., the stellar parameters) for a new star outside the training set in a
-fraction of a second, which is about five orders of magnitude faster than
-existing analysis codes. We refer the reader to \citet{Ness15a} for a more
-detailed description of \tc{}.
+(i.e., the stellar parameters) for a new star outside the training set. We refer
+the reader to \citet{Ness15a} for a more detailed description of \tc{}.
 
@@
 expect equally impressive results.  Some chemical abundances only affect few
-pixels of the spectrum. Other elements only present transitions that are
-frequently blended with other elements, implying covariance between elemental
-abundances.  And all transitions will have linear (or higher-order) 
-relationships with stellar parameters.  Addressing these issues by increasing
-the complexity of the label vector introduces additional issues (e.g.,
-over-fitting, interpretability).  Although the existing \textit{Cannon}
-framework can be extended to include individual chemical abundances, they cannot
-be introduced by extensions of the label vector without additional issues.
+pixels of the spectrum. Other species (element and ionisation stage) only
+experience stimulated emission at wavelengths that are frequently blended
+with other species, inducing covariance between elemental abundances.
+And all transitions will likely have linear (or higher-order) relationships with
+stellar parameters.  Addressing these issues by increasing the complexity 
+of the label vector introduces additional issues (e.g., over-fitting, interpretability).
+Although the existing \textit{Cannon} framework can be extended to include
+individual chemical abundances, they cannot be introduced by extensions 
+of the label vector without considering these additional issues.
 
-% TODO: rewrite this so it all flows better.
+\subsection{Using the behaviour of spectral lines to understand why \tc\ works}
+
 Before we extend \tc{} to include individual chemical abundances, it is a useful
@@
 line is well-represented by a Gaussian profile at this spectral resolution, the 
-flux at $\lambda_r$ is related to the EW of the line through
+flux at $\lambda_r$ is related to the EW of the line in star $n$ ($\mathcal{W}_n$):
 
 \begin{equation}
-\mathcal{W} = \sigma{}\sqrt{2\pi}(1 - f_{n\lambda_r})
+\mathcal{W}_n = \sigma{}\sqrt{2\pi}(1 - f_{n\lambda_r}).
 \label{eq:ew-flux}
@@
 
-where $\sigma$ is the standard deviation of the profile. For typical FGK-type
-stars observed with $\mathcal{R} \lesssim 20,000$, $\sigma$ will be dominated by
-the line spread function (LSF). If we presume that the LSF is well-known, or 
-reasonably approximated by a fixed spectral resolution $\mathcal{R}$, then it is
-clear to see that \tc{} is implicitly modelling the EW of an atomic line as a 
-function of stellar labels, and modelling spectral flux through a per-pixel 
-scaling of $\sigma$. Indeed, the success of \tc{} suggests that the EW of an 
-atomic line can also be well-approximated by a simple model relating the labels 
-(i.e., stellar parameters and abundance of a single species). If such a model 
-could be constructed and appropriately included within the existing 
-\textit{Cannon} framework, it would permit the determination of individual 
-chemical abundances, whilst maintaining the flexibility in training on labels 
-with poorly-known relationships on spectral fluxes (e.g., ages).
+Here $\sigma$ is the standard deviation of the profile. For typical FGK-type
+stars observed with $\mathcal{R} \lesssim 20,000$, $\sigma$ will be dominated
+by the line spread function (LSF) of the spectrograph. Because \tc{} is best suited
+to large spectroscopic surveys, we can presume that the LSF is well-known, or
+at least reasonably constant for the training set, perhaps approximated by the 
+quoted spectral resolution $\mathcal{R}$. In Section \ref{sec:something} we use
+the knowledge of the spectrograph LSF, but for explanatory purposes here we will
+assume it is constant for all $n$ stars surrounding the profile at $\lambda_r$.
+
+If the profile width is constant, then it is straightforward to see by integrating
+\tc\ model the entire profile, \tc\ is implicitly modelling the \ew\ of an atomic line
+as a function of stellar labels, and modelling spectral flux through a per-pixel
+scaling of $\sigma_\lambda$. Indeed, the success of \tc\ implies that the 
+strength \ew\ of an atomic line could be well-approximated by a quadratic
+model of a label vector containing the stellar parameters, where \mh\ is 
+actually a proxy for the abundance of the atomic line in question. If such a
+\textit{line strength model} could be constructed and appropriately
+included within the existing \textit{Cannon} framework, it would permit the 
+determination of individual chemical abundances, whilst maintaining the 
+flexibility in training on labels with poorly-known relationships on spectral 
+fluxes (e.g., ages).
 
@@
 line are still well-modelled by \tc{} (as judged by $\ssq$ and performance in
-cross-validation tests). Indeed, the line $\mathcal{W}$ would still be 
+cross-validation tests). Indeed, the line strength \ew\ would still be 
 predictable with high-fidelity, implying that the coefficients $\cv$ for pixels
-in the wings just scale in a different way. Equation \ref{eq:ew-flux} allows us
-to directly translate $\mathcal{W}$ to $f_{n\lambda}$, thereby accounting for 
-the flux contribution from a single atomic line. If the spectral resolving power
-were sufficiently high such that the absorption profiles appeared Lorentzian in
-nature, some prior information on the shape $\gamma$ would be required to 
-translate $\mathcal{W}$ to $f_{n\lambda}$. Thus, while Equation \ref{eq:ew-flux}
-allows the translation from $\mathcal{W}$ to $f_{n\lambda}$ for 
-$\mathcal{R} \lesssim 20,000$ spectra, given some knowledge of the LSF, no 
-limiting interpretation about \tc{} should be drawn. Importantly, all future 
+in the wings just scale in a different way.
+
+The crucial inference from Equation \ref{eq:ew-flux} is that knowing the LSF
+allows us to directly translate \ew\ to $f_{n\lambda}$, thereby accounting for 
+the flux contribution from a single atomic line across all neighbouring pixels.
+If the spectral resolving power were sufficiently high such that the absorption
+profiles appeared Lorentzian, some prior information on the shape $\gamma$
+may be required to translate \ew\ to $f_{n\lambda}$. Importantly, all future 
 stellar spectroscopic surveys are currently planning for spectral resolutions
@@
 extension of \tc{}.
+
+To demonstrate that \tc\ is implicitly modelling the EW of an atomic line as
+a function of stellar parameters and the abundance of the given element,
+here we employ the quadratic-in-labels model from \citet{Ness2015a} to
+model the strength of a single atomic line. We used the NNN stars from the
+training set used by \citet{Ness2015a} and calculated the abundance of a
+Si 1 line at N \AA\ for thirty different EWs uniformly sampled between 5~m\AA\
+and 120~m\AA. These stars cover the main sequence and red giant branch
+(RGB), and the overall metallicity spans three decades from -2.5 to +0.5 dex.
+The abundances were calculated using MARCS model atmospheres \citep{marcs}
+and a modified Cython version of MOOG \citep[][2014 version]{Sneden1973}
+which includes ABO broadening of atomic lines \citep{ABO} and several 
+improvements to the treatment of hydrogen opacities. The atomic data
+employed is listed in Table \ref{tab:atomic-data}. We describe the 
+\textit{line strength model} that relates the line strength \ew\ for the $m$-th
+atomic line in star $n$ as,
+
+\begin{equation}
+\mathcal{W}_{n,m} \approx \psiv^{\intercal}\cdot\lv
+\label{eq:line-strength-model}
+\end{equation}
+
+\noindent{}where the label vector 
+$\lv = \{T_{\rm eff},\log{g},\log_{\epsilon}{\rm Si}\}$\footnote{Recall
+${\rm [Si/Fe]} \equiv {\rm [Si/H]} - {\rm [Fe/H]} \equiv \log_{\epsilon}{\rm Si} 
+- \log_{\epsilon}{\rm Si}_\odot - {\rm [Fe/H]}$, where we adopt the \citet{Asplund2009}
+solar abundances.} and the coefficients of the vector $\psiv$ for a given 
+atomic line need to be solved for. Combining Equations \ref{eq:ew-flux} and
+\ref{eq:line-strength-model} yields the approximate flux $f_{n,\lambda_r}$ that 
+atomic line $m$ contributes to star $n$ at wavelength $\lambda_r$, which we
+describe as $h(\lv|\psiv,\sigma)$:
+
+\begin{equation}
+f_{n,\lambda_r} \approx h(\lv|\psiv,\sigma,\lambda) \equiv 1 - \frac{\psiv^{\intercal}\cdot\lv}{\sigma\sqrt{2\pi}}
+\end{equation}
+
+We calculate $\psiv$ given the stellar parameters of the training set and the
+calculated $\logeps{(Si~1)}$ abundances from MOOG, calculated at many \ew\ values.
+The difference between the MOOG-calculated abundance and our line strength model
+is shown in the second panel of Figure \ref{fig:atomic-line-model-example}.
+Negligible bias is present, and the standard deviation is 0.01~dex, demonstrating
+that a quadratic-in-labels model can relate line strength \ew\ and abundance
+for all types of stars in the training set. This is further shown in the third panel of 
+Figure \ref{fig:atomic-line-model-example},
+which shows the spectrum surrounding the Si 1 line for a random star\footnote{The
+star is truly chosen at random, see Appendix \ref{appendix:reproducibility} for steps
+to reproduce this research.} in the training set. We calculated the corresponding
+line strength \ew\ given the stellar parameters and reported Si 1 abundance, 
+adopted a spectral resolution of $\mathcal{R} = 22,500$ and produced a model 
+spectrum of this atomic line following Equation \ref{ew-flux}.
+
+We note that some of the 0.01~dex scatter between our line strength model and 
+MOOG (second panel of Figure \ref{fig:atomic-line-model-example}) is likely due to 
+the tolerance employed by MOOG's optimisation algorithm to calculate a line 
+abundance, given an EW.
+
+% Show a figure with different EW vs abundances, colored by teff/logg
+% Then show the expected EW vs synthesis EW, and show the residual
+% Show a portion of the spectrum where we translate a modelled EW into a spectrum
+% by using some knowledge of the LSF
+
+% h relates W to flux values.
+% W
+
+% Discuss using the Cannon model from Ness et al to model the EW as a func. of stellar params.
+
+\subsection{Using the behaviour of spectral lines to guide model complexity}
+
+Knowing that a simple model can relate the a single line abundance 
+$\log_{\epsilon}{\rm X}$ to the equivalent width \ew\, given the stellar
+parameters, it is intriguing to ask whether model atmosphere relations and
+the theory of spectral line formation can guide the level of model complexity
+required.
 
@@
 
-Knowing that a simple model can relate the a single line abundance [X/H] and integrated strength of an atomic line $\mathcal{W}$, given stellar parameter labels, it is intriguing to ask whether model atmosphere relations and the theory of spectral line formation can guide the level of model complexity required.
+% How does EW of a line change w.r.t. stellar parameters
+\textbf{DESCRIPTION OF HOW EW FOR A SINGLE LINE WILL CHANGE WITH RESPECT TO STELLAR PARAMETERS AND ABUNDANCE FOR A NEUTRAL LINE IN A GRAY ATMOSPHERE.}
 
-% How does EW of a line change w.r.t. stellar parameters
-\textbf{Description of how EW for a single line will change w.r.t stellar parameters}
+\subsection{Extending \tc{} to include individual atomic lines}
 
@@
 \begin{thebibliography}{99}
+
+\bibitem[Asplund et al.(2009)]{Asplund2009} Asplund, M., Grevesse, N., Sauval, A.~J., \& Scott, P.\ 2009, \araa, 47, 481
+
 \bibitem[Astropy Collaboration et al.(2013)]{astropy} Astropy Collaboration, Robitaille, T.~P., Tollerud, E.~J., et al.\ 2013, \aap, 558, AA33
@@
 
-\section{Appendix A: Reproducing this research}
+\appendix
+\section{Reproducing this work}
+\label{appendix:reproducibility}
+
 This \article{} has been compiled from the \texttt{git} repository (abbreviated commit hash \texttt{\gitAbbrevHash} on \texttt{\gitBranch} branch) hosted online by GitHub at \texttt{github.com/andycasey/fireworks}.

### 2015-10-27 article/ms.tex
Things

@@
 \usepackage{gitinfo2}
+
+\long\def\/*#1*/{}
 
@@
 \begin{equation}
-\mathcal{W}_n = \sigma{}\sqrt{2\pi}(1 - f_{n\lambda_r}).
+\mathcal{W}_n = \sigma{}a\sqrt{2\pi}.
 \label{eq:ew-flux}
@@
 
-Here $\sigma$ is the standard deviation of the profile. For typical FGK-type
+Here $a$ is the amplitude of the profile ($a = 1 - f_{n\lambda_r}$) and $\sigma$ 
+is the standard deviation. For typical FGK-type
 stars observed with $\mathcal{R} \lesssim 20,000$, $\sigma$ will be dominated
@@
 solar abundances.} and the coefficients of the vector $\psiv$ for a given 
-atomic line need to be solved for. Combining Equations \ref{eq:ew-flux} and
-\ref{eq:line-strength-model} yields the approximate flux $f_{n,\lambda_r}$ that 
-atomic line $m$ contributes to star $n$ at wavelength $\lambda_r$, which we
-describe as $h(\lv|\psiv,\sigma)$:
+atomic line need to be solved for. Thus from Equations \ref{eq:ew-flux} and
+\ref{eq:line-strength-model} we obtain the approximate flux $f_{n,\lambda_r}$ that 
+atomic line $m$ (at wavelength $\lambda_m$) contributes to star $n$ at any
+wavelength $\lambda$:
 
 \begin{equation}
-f_{n,\lambda_r} \approx h(\lv|\psiv,\sigma,\lambda) \equiv 1 - \frac{\psiv^{\intercal}\cdot\lv}{\sigma\sqrt{2\pi}}
+f_{n,\lambda_r} \approx 1 - \left(\psiv^{\intercal}\cdot\lv\right)\mathcal{N}\left(\lambda_{m} - \lambda_r,\sigma\right).
 \end{equation}
@@
 We calculate $\psiv$ given the stellar parameters of the training set and the
-calculated $\logeps{(Si~1)}$ abundances from MOOG, calculated at many \ew\ values.
+calculated $\logeps{(Si~1)}$ abundances from MOOG, calculated for multiple \ew.
 The difference between the MOOG-calculated abundance and our line strength model
 is shown in the second panel of Figure \ref{fig:atomic-line-model-example}.
-Negligible bias is present, and the standard deviation is 0.01~dex, demonstrating
+The agreement is excellent. Negligible bias is present, and the standard deviation is 0.01~dex, demonstrating
 that a quadratic-in-labels model can relate line strength \ew\ and abundance
@@
 adopted a spectral resolution of $\mathcal{R} = 22,500$ and produced a model 
-spectrum of this atomic line following Equation \ref{ew-flux}.
-
-We note that some of the 0.01~dex scatter between our line strength model and 
-MOOG (second panel of Figure \ref{fig:atomic-line-model-example}) is likely due to 
-the tolerance employed by MOOG's optimisation algorithm to calculate a line 
-abundance, given an EW.
+spectrum of this atomic line following Equation \ref{ew-flux}. We note that some 
+of the 0.01~dex scatter between our 
+line strength model and MOOG (second panel of Figure 
+\ref{fig:atomic-line-model-example}) is likely due to the tolerance employed by 
+MOOG's optimisation algorithm to calculate a line abundance, given an EW.
 
@@
 
-% h relates W to flux values.
-% W
-
-% Discuss using the Cannon model from Ness et al to model the EW as a func. of stellar params.
-
 \subsection{Using the behaviour of spectral lines to guide model complexity}
 
-Knowing that a simple model can relate the a single line abundance 
-$\log_{\epsilon}{\rm X}$ to the equivalent width \ew\, given the stellar
+Knowing that a quadratic model can relate the a single line abundance 
+$\logeps{X}$ to the equivalent width \ew\, given the stellar
 parameters, it is intriguing to ask whether model atmosphere relations and
@@
 
-We now have a functional form of a model that relates $\mathcal{W}$ to $\log_\epsilon\left({\textrm X}\right)$ (or a scaled-solar [X/H] value), given the stellar parameters, which themselves are part of the set of $K$ labels.
-It is interesting to note that Equation \ref{eq:ew-logx-model} is not dissimilar to the quadratic-in-labels model used by \citet{Ness15a} for APOGEE data.
-With only the prior knowledge that \textit{a simple model of the labels can predict spectral fluxes}, we have derived the appropriate functional form for a single absorption line.
-The model $\mathcal{W}\left({\rm [X/H]}, T, logg\right)$ simply approximates the curve-of-growth for a single line, and the coefficients still need to be calculated.
-The practicalities of determining these coefficients is outlined in Section \ref{sec:somewhere}.
-For the purposes of introducing the model, we can assume that they are easily calculable.
+We now have a physically-motivated model that relates \ew\ to $\logeps{X}$
+(or a scaled-solar [X/H] or [X/Fe] ratio), given the stellar parameters. Here we
+describe how we include multiple atomic line models as part of \tc's training
+and testing phases.
 
-Consider multiple atomic lines, where the coefficients to translate $\mathcal{W}$ and [X/H] are known.
-Given some knowledge of the line spread function, the EWs $\mathcal{W}$ to flux values $f_{n\lambda}$ can be translated such that:
+% different psi_m for each line at lamgda_m
+
+% Introduce a h function that includes treatment from all lines.
 
 \begin{equation}
-h = X
-\label{eq:h-model}
+h(\lv\given\{\psiv\},\{\lambda_m\},\sigma) = 1 - \prod^{M}_{m=0}(\psiv^\intercal\cdot\lv)\mathcal{N}(\lambda_m - \lambda,\sigma)
 \end{equation}
 
-Here we implicitly assume the mutual opacity contributions of nearby weak lines will be small. 
-At this point it is worthwhile to ask whether Equation \ref{eq:h-model} can be scaled sufficiently to $M$ atomic lines in order to produce an entire spectral model for $f_{n\lambda}$, without the contribution of $g\left(\lv\given\cv\right)$.
-There would be a number of immediate issues with that approach: a huge number of additional coefficients would be required, hidden lines could not be encapsulated accurately by the model, strong lines would be extremely poorly modelled by the fit, no longer linear (slow), but most importantly the pixels would lose flexibility in training on arbitrary labels.
-
-If lines are blended, one could consider modelling many more atomic transitions.
-However as the number grows, this situation becomes unnecessarily complex:
-often blended lines exist and we do not care about their abundance, only their
-inclusion (nuisance or dirty lines). Other features may be a complex combination
-of thousands of molecular features.
-
-Here we introduce a method of including individual atomic transitions within an existing \textit{Cannon} model.
-
-The spectral model becomes a multiplicative function of $g$ and $h$ with some
-unknown noise.
+% Assumptions for the training set (complete abundances). Incomplete
+% abundances in the training set can be discussed in Section ref.
 
@@
 
-This inclusion combines the existing flexibility in the model, allowing for training on arbitrary labels, and uses individual absorption lines for which some information is known (species, atomic data, etc) in order to trace the chemical abundances of individual elements.
+The training stage can be described in practice as the following:
 
-The vector $\wv$ can be calculated directly from spectral synthesis, or using data.
-In practice, given some rest-frame normalised spectra and some knowledge of the
-line spread function, $h(\lv,\wv\given{}X)$ can be calculated before training \tc.
+Build all the atomic lines of interest.
 
-%Utilising a simple polynomial model to predict spectral fluxes is a distinctive
-%method from parallel efforts in radiative transfer of realistic (3D) stellar 
-%photospheres. Calculations of this nature are extremely expensive. Indeed, even
-%with simplifications of these models (e.g., interpolated 1D photospheres),
-%calculating synthetic spectra requires knowledge of species opacities, detailed 
-%atomic and molecular information for all contributing elements, solving for
-%the equation of state, and iteratively performing expensive radiative transfer
-%calculations. Given the
-%simplicity of a quadratic-in-labels spectral model with \tc{}, it is remarkable
-%that these complex calculations can be simplified so greatly, whilst maintaining
-%the typical measurement fidelity reported for high-resolution spectroscopic studies based on
-%\textit{curve-of-growth} analyses.
+For each star
+\begin{enumerate}
+\item Use the reference abundance for each elemental species to calculate expected EWs
+\item Use the LSF to calculate $h$
+\item Divide the flux by $h$
+\end{enumerate}
 
-% Figure of relating EW to [X/H] with some simple model compared to analytic value.
+For each pixel
+\begin{enumerate}
+\item Train \tc{} on the residuals
+\end{enumerate}
 
-The compromise is to use TC and `clean' lines that we care about in concert.
+This framework combines the existing flexibility in \tc\, allowing for training
+on arbitrary labels, and uses individual absorption lines for which some 
+information is known (species, atomic data, etc) in order to trace the chemical 
+abundances of individual elements.
 
-% existing simple model. replaces expensive synthesis, involving model atmosphere
-% generation, t-tau relations, opacities, equation of state, integrated line-of-sight
-% and radiative transfer, etc.
-
-% before extending this model to include individual abundances, it is worthwhile
-% demonstrating (from a physics perspective) *why* the cannon works so well.
-
-% APOGEE, R of 20,000, 
-% LAMOST, higher Rs
-% super high-resolution also works.
+At testing time, the story is a little different. Because the stellar parameters
+influence both $h$ and $g$, we must solve for the stellar parameters and all
+chemical abundances simultaneously.
 
 \section{APOGEE}
+
+We will use the APOGEE X sample to derive individual chemical abundances of 15
+elements using \tc{}.
 
@@
 % Limitations: what about missing abundance data?
+
+\/* 
+% for the discussion.
+Here we implicitly assume the mutual opacity contributions of nearby weak lines will be small. 
+At this point it is worthwhile to ask whether Equation \ref{eq:h-model} can be scaled sufficiently to $M$ atomic lines in order to produce an entire spectral model for $f_{n\lambda}$, without the contribution of $g\left(\lv\given\cv\right)$.
+There would be a number of immediate issues with that approach: a huge number of additional coefficients would be required, hidden lines could not be encapsulated accurately by the model, strong lines would be extremely poorly modelled by the fit, no longer linear (slow), but most importantly the pixels would lose flexibility in training on arbitrary labels.
+
+If lines are blended, one could consider modelling many more atomic transitions.
+However as the number grows, this situation becomes unnecessarily complex:
+often blended lines exist and we do not care about their abundance, only their
+inclusion (nuisance or dirty lines). Other features may be a complex combination
+of thousands of molecular features.
+*/
 

### 2015-10-28 fireworks/tests/test_utils.py
Added element/species converters

@@ -1,0 +1,57 @@
+#!/usr/bin/env python
+# -*- coding: utf-8 -*-
+
+""" Test the general utilities. """
+
+__author__ = "Andy Casey <arc@ast.cam.ac.uk>"
+
+import unittest
+
+from fireworks import utils
+
+class TestElementRepresentations(unittest.TestCase):
+    """
+    Test transformations between element and ionisation stage.
+    """
+
+    def setUp(self):
+
+        table =       '''H                                                  He
+                         Li Be                               B  C  N  O  F  Ne
+                         Na Mg                               Al Si P  S  Cl Ar
+                         K  Ca Sc Ti V  Cr Mn Fe Co Ni Cu Zn Ga Ge As Se Br Kr
+                         Rb Sr Y  Zr Nb Mo Tc Ru Rh Pd Ag Cd In Sn Sb Te I  Xe
+                         Cs Ba Lu Hf Ta W  Re Os Ir Pt Au Hg Tl Pb Bi Po At Rn
+                         Fr Ra Lr Rf'''
+
+        lanthanoids = '''La Ce Pr Nd Pm Sm Eu Gd Tb Dy Ho Er Tm Yb'''
+        actinoids   = '''Ac Th Pa U  Np Pu Am Cm Bk Cf Es Fm Md No'''
+
+        self.periodic_table = table.replace(' Ba ', ' Ba ' + lanthanoids + ' ')\
+            .replace(' Ra ', ' Ra ' + actinoids + ' ').split()
+        
+
+    def test_get_species(self):
+        for i, element in enumerate(self.periodic_table, start=1):
+            self.assertEqual(i, utils.species(element))
+
+
+    def test_get_species_any_casing(self):
+        for i, element in enumerate(self.periodic_table, start=1):
+            self.assertEqual(i, utils.species(element.upper()))
+            self.assertEqual(i, utils.species(element.lower()))
+
+
+    def test_get_element(self):
+        for i, element in enumerate(self.periodic_table, start=1):
+            self.assertEqual(element, utils.element(i))  
+
+
+    def test_get_ionisation(self):
+        for i, element in enumerate(self.periodic_table, start=1):
+            for j in range(1, 3):
+                self.assertEqual((element, j), 
+                    utils.element(i + (j - 1) * 0.1, True))
+
+
+

### 2015-10-28 fireworks/utils.py

@@ -13,6 +13,110 @@ from itertools import combinations_with_replacement
 
 logger = logging.getLogger("cannon")
 
+
+def element(species, full_output=False):
+    """
+    Convert a floating point representation of a species to a string
+    representation of the element, and optionally, its ionisation level.
+    
+    :param species:
+        The floating-point representation of the transition.
+
+    :type species:
+        float or int
+
+    :param full_output: [optional]
+        Return the ionisation stage as well as the element.
+
+    :type full_output:
+        bool
+
+    :returns:
+        The element name, and optionally the ionisation level.
+    """
+    
+    periodic_table = '''H                                                  He
+                        Li Be                               B  C  N  O  F  Ne
+                        Na Mg                               Al Si P  S  Cl Ar
+                        K  Ca Sc Ti V  Cr Mn Fe Co Ni Cu Zn Ga Ge As Se Br Kr
+                        Rb Sr Y  Zr Nb Mo Tc Ru Rh Pd Ag Cd In Sn Sb Te I  Xe
+                        Cs Ba Lu Hf Ta W  Re Os Ir Pt Au Hg Tl Pb Bi Po At Rn
+                        Fr Ra Lr Rf'''
+    
+    lanthanoids    = '''La Ce Pr Nd Pm Sm Eu Gd Tb Dy Ho Er Tm Yb'''
+    actinoids      = '''Ac Th Pa U  Np Pu Am Cm Bk Cf Es Fm Md No'''
+    
+    periodic_table = periodic_table.replace(' Ba ', ' Ba ' + lanthanoids + ' ')\
+        .replace(' Ra ', ' Ra ' + actinoids + ' ').split()
+    del actinoids, lanthanoids
+
+    try:
+        species = float(species)
+    except (TypeError, ValueError):
+        raise TypeError("species must be represented by a floating point-type")
+
+    if int(species) > len(periodic_table):
+        raise ValueError("""species '{0}' not recognised 
+            (molecular representations not supported)""".format(species))
+        
+    element = periodic_table[int(species) - 1]
+    ionisation = int(round(10 * (species % int(species))) + 1)
+    
+    return (element, ionisation) if full_output else element
+
+
+def species(element):
+    """
+    Converts a string representation of an element and its ionisation state
+    to a floating species value.
+
+    :param element:
[…]

### 2015-10-28 fireworks/tests/test_utils.py
Updated utils

@@ -36,10 +36,35 @@ class TestElementRepresentations(unittest.TestCase):
             self.assertEqual(i, utils.species(element))
 
 
-    def test_get_species_any_casing(self):
+    def test_get_species_lowercase(self):
+        for i, element in enumerate(self.periodic_table, start=1):
+            self.assertEqual(i, utils.species(element.lower()))
+
+
+    def test_get_species_uppercase(self):
         for i, element in enumerate(self.periodic_table, start=1):
             self.assertEqual(i, utils.species(element.upper()))
-            self.assertEqual(i, utils.species(element.lower()))
+            
+
+    def test_get_species_roman(self):
+        for i, element in enumerate(self.periodic_table, start=1):
+            for j in range(3):
+                self.assertEqual(i + 0.1 * j,
+                    utils.species("{0} {1}".format(element, "I" * (j + 1))))
+
+
+    def test_get_species_roman_lowercase(self):
+        for i, element in enumerate(self.periodic_table, start=1):
+            for j in range(3):
+                self.assertEqual(i + 0.1 * j,
+                    utils.species("{0} {1}".format(element.lower(), "i" * (j + 1))))
+
+
+    def test_get_species_roman_uppercase(self):
+        for i, element in enumerate(self.periodic_table, start=1):
+            for j in range(3):
+                self.assertEqual(i + 0.1 * j,
+                    utils.species("{0} {1}".format(element.upper(), "I" * (j + 1))))
 
 
     def test_get_element(self):
@@ -52,6 +77,4 @@ class TestElementRepresentations(unittest.TestCase):
             for j in range(1, 3):
                 self.assertEqual((element, j), 
                     utils.element(i + (j - 1) * 0.1, True))
-
-
 

### 2015-10-28 fireworks/utils.py

@@ -7,6 +7,7 @@ __author__ = "Andy Casey <arc@ast.cam.ac.uk>"
 
 import sys
 import logging
+import numpy as np
 from time import time
 from collections import Counter
 from itertools import combinations_with_replacement
@@ -113,8 +114,51 @@ def species(element):
     else:
         ionisation = 0
 
-    return 1 + periodic_table.index(element) + ionisation/0.1
+    return 1 + periodic_table.index(element) + ionisation*0.1
     
+
+def xi_relation(effective_temperature, surface_gravity):
+    """
+    Estimate microtubulence from relations between effective temperature and
+    surface gravity. For giants (logg < 3.5) the relationship employed is from
+    Kirby et al. (2008, ) and for dwarfs (logg >= 3.5) the Reddy et al. (2003)
+    relation is used.
+
+    :param effective_temperature:
+        The effective temperature of the star in Kelvin.
+
+    :type effective_temperature:
+        float
+
+    :param surface_gravity:
+        The surface gravity of the star.
+
+    :type surface_gravity:
+        float
+
+    :returns:
+        The estimated microturbulence (km/s) from the given stellar parameters.
+
+    :rtype:
+        float
+    """
+
+    try:
+        _ = len(effective_temperature)
+
+    except TypeError:
+        if surface_gravity >= 3.5:
+            xi = 1.28 + 3.3e-4 * (effective_temperature - 6000) \
+                - 0.64 * (surface_gravity - 4.5)
+        else:
+            xi = 2.70 - 0.509 * surface_gravity
+    else:
+        xi = np.nan * np.ones(len(effective_temperature))
+        dwarfs = surface_gravity >= 3.5
+        xi[dwarfs] = 1.28 + 3.3e-4 * (effective_temperature[dwarfs] - 6000) \
+            - 0.64 * (surface_gravity[dwarfs] - 4.5)
+        xi[~dwarfs] = 2.70 - 0.509 * surface_gravity[~dwarfs]
+    return xi
 
 
 def progressbar(iterable, message=None, size=100):

### 2015-10-28 fireworks/atomic.py
Refactoring atomic line stuff

@@ -8,11 +8,371 @@ __author__ = "Andy Casey <arc@ast.cam.ac.uk>"
 import logging
 import numpy as np
 import scipy.optimize as op
+from astropy.table import Table
 
 from oracle import photospheres, synthesis
 
 logger = logging.getLogger("fireworks")
 
+import utils
+
+def _guess_abundance_label_format(species, star_labels):
+    """
+    Guess the string format of the abundance label columns, based on the atomic
+    species we expect to be measureable and the existing labels for each star.
+
+    :param species:
+        A list of floating-point representations of transitions.
+
+    :type species:
+        iterable
+
+    :param star_labels:
+        The list of available string labels for each star.
+
+    :type star_labels:
+        list of str
+
+    :returns:
+        A two-length tuple containing the guessed string format of the abundance
+        label, and what each element is measured relative to in the Sun (H, Fe,
+        or None for logarithmic abundances).
+    """
+
+    try_these = [
+        ("{ELEMENT}_H", "H"),
+        ("{Element}_H", "H"),
+        ("{ELEMENT}{ion}", None),
+        ("{Element}{ion}", None),
+        ("{ELEMENT}_FE", "Fe"),
+        ("{Element}_Fe", "Fe"),
+        ("LOG_{ELEMENT}", None),
+        ("LOG_{ELEMENT}{ion}", None),
+        ("logeps_{element}{ion}", None),
+    ]
+    if not len(species): return try_these[0]
+
+    for element_repr, relative_to in try_these:
+        for each in set(species):
+            element, ion = utils.element(each, True)
+            dict_repr = {
+                "ELEMENT": element.upper(),
+                "Element": element.title(),
+                "element": element.lower(),
+                "ion": ion
+            }
+            if element_repr.format(**dict_repr) not in star_labels:
+                break
+        else:
+            break
+    else:
+        raise ValueError("Could not guess label abundance format. Tried: {0}"\
+            .format(", ".join([_[0] for _ in try_these])))
+
+    return (element_repr, relative_to)
+
+
+
+
+
+class AtomicTransitionModel(object):
+
+
+    def __init__(self, stars, atomic_data, abundance_label_format=None,
+        stellar_parameter_labels=None):
+        """
+        Initialise the atomic transition model.
+        """
+
+        self.stars = stars
+        self.atomic_data = atomic_data
+
+        print("CHECK THE FORMAT OF STARS AND ATOMIC DATA")
+
[…]

### 2015-10-29 fireworks/atomic.py
Introduce AtomicLineModel

@@ -5,8 +5,10 @@
 
 __author__ = "Andy Casey <arc@ast.cam.ac.uk>"
 
+import cPickle as pickle
 import logging
 import numpy as np
+import os
 import scipy.optimize as op
 from astropy.table import Table
 
@@ -14,7 +16,7 @@ from oracle import photospheres, synthesis
 
 logger = logging.getLogger("fireworks")
 
-import utils
+import model, utils
 
 def _guess_abundance_label_format(species, star_labels):
     """
@@ -74,265 +76,235 @@ def _guess_abundance_label_format(species, star_labels):
 
 
 
+class MetaAtomicTransition(type):
 
-class AtomicTransitionModel(object):
+    """
+    Meta class to enforce read-only properties on the AtomicTransitionModel.
+    """
+
+    @property
+    def element(cls):
+        return cls.element
+
+    @property
+    def ion(cls):
+        return cls.ion
+
+    @property
+    def wavelength(cls):
+        return cls.wavelength
+
+    @property
+    def loggf(cls):
+        return cls.loggf
+
+    @property
+    def excitation_potential(cls):
+        return cls.excitation_potential
+
+    @property
+    def solar_abundance(cls):
+        return cls.solar_abundance
 
 
-    def __init__(self, stars, atomic_data, abundance_label_format=None,
-        stellar_parameter_labels=None):
+class AtomicTransition(object):
+
+    __metaclass__ = MetaAtomicTransition
+
+    def __init__(self, data):
         """
-        Initialise the atomic transition model.
+        Initialise an AtomicTransition object.
+
+        :param data:
+            The relevant atomic data for this transition. This must include at
+            least the wavelength, species, excitation potential, and the 
+            oscillator strength (as loggf).
+
+        :type data:
+            dict or :class:`~astropy.table.Row`
+
         """
 
-        self.stars = stars
-        self.atomic_data = atomic_data
+        # Require species or element/ion.
+        try:
+            species = data["species"]
 
-        print("CHECK THE FORMAT OF STARS AND ATOMIC DATA")
+        except (TypeError, KeyError):
+            try:
+                element, ion = data["element"], data["ion"]
 
-        self.abundance_label_format = abundance_label_format \
-            or _guess_abundance_label_format(
-                atomic_data["species"], stars.dtype.names)
+            except (TypeError, KeyError):
+                raise KeyError(
+                    "must provide either species or (element and ion)")
 
[…]

### 2015-10-29 fireworks/model.py

@@ -15,7 +15,7 @@ def requires_training_wheels(f):
     """
     def wrapper(model, *args, **kwargs):
         if not model._trained:
-            raise TypeError("the model needs training first")
+            raise TypeError("the model needs training/optimising first")
         return f(model, *args, **kwargs)
     return wrapper
 
